Heterozygosity for neurodevelopmental disorder-associated TRIO variants yields distinct deficits in behavior, neuronal development, and synaptic transmission in mice

Version 1

Abstract

Abstract

Genetic variants in TRIO are associated with neurodevelopmental disorders (NDDs) including schizophrenia (SCZ), autism spectrum disorder (ASD) and intellectual disability. TRIO uses its two guanine nucleotide exchange factor (GEF) domains to activate GTPases (GEF1: Rac1 and RhoG; GEF2: RhoA) that control neuronal development and connectivity. It remains unclear how discrete TRIO variants differentially impact these neurodevelopmental events. Here, we investigate how heterozygosity for NDD-associated Trio variants – +/K1431M (ASD), +/K1918X (SCZ), and +/M2145T (bipolar disorder, BPD) – impact mouse behavior, brain development, and synapse structure and function. Heterozygosity for different Trio variants impacts motor, social, and cognitive behaviors in distinct ways that align with clinical phenotypes in humans. Trio variants differentially impact head and brain size with corresponding changes in dendritic arbors of motor cortex layer 5 pyramidal neurons (M1 L5 PNs). Although neuronal structure was only modestly altered in the Trio variant heterozygotes, we observe significant changes in synaptic function and plasticity. We also identified distinct changes in glutamate synaptic release in +/K1431M and +/M2145T cortico-cortical synapses. The TRIO K1431M GEF1 domain has impaired ability to promote GTP exchange on Rac1, but +/K1431M mice exhibit increased Rac1 activity, associated with increased levels of the Rac1 GEF Tiam1. Acute Rac1 inhibition with NSC23766 rescued glutamate release deficits in +/K1431M variant cortex. Our work reveals that discrete NDD-associated Trio variants yield overlapping but distinct phenotypes in mice, demonstrates an essential role for Trio in presynaptic glutamate release, and underscores the importance of studying the impact of variant heterozygosity in vivo.

Introduction

Neurodevelopmental disorders (NDDs), including autism spectrum disorder (ASD), intellectual disability (ID), developmental delay (DD), schizophrenia (SCZ) and bipolar disorder (BPD), disrupt brain development and function. NDDs share considerable comorbidities and present with overlapping symptoms, including impairments in cognition, behavior, language, social and motor skills, emotions, and learning ability (1). In addition, the risk genes for NDDs overlap significantly, further suggesting shared mechanistic underpinnings and pathology for these disorders (2, 3).

TRIO encodes a large cytoskeletal regulatory protein with two guanine nucleotide exchange factor (GEF) domains for Rho family GTPases - GEF1 activates Rac1 and RhoG, while GEF2 activates RhoA (4–6). TRIO relays signals from cell surface receptors, acting on GTPases to coordinate cytoskeletal rearrangements critical for proper neurodevelopment (7–15). Trio knockout mice exhibit decreased survival, skeletal muscle defects, and severe defects in brain development (16–18). Selective ablation of Trio in either excitatory or inhibitory neurons alters their morphology, impairs synaptic signaling, and yields NDD-related behavioral defects (19, 20). Trio deficiency also leads to aberrations in long-term potentiation (LTP), as Trio-Rac1 signaling promotes AMPA receptor trafficking to increase synaptic strength (19, 21, 22).

Damaging de novo mutations and ultra-rare variants in TRIO are enriched in individuals with NDDs (23–30). Interestingly, nonsense variants spread throughout TRIO are enriched in individuals with SCZ (30, 31), whereas pathogenic missense TRIO variants in or surrounding the GEF1 domain are associated with ASD/ID (24, 28, 29). Variants in the Trio GEF1 domain that decrease Rac1 activity are associated with milder ID and microcephaly, whereas variants in the adjacent spectrin repeat 8 domain that increase Rac1 activity are associated with more severe ID and macrocephaly (25, 26, 32, 33). Rare missense variants in TRIO have also been observed in BPD, epilepsy, and other disorders, but studies to date are underpowered to establish a causal link with these disorders. How distinct heterozygous TRIO variants differentially impact mammalian brain development and lead to diverse phenotypes remains a fundamental and unresolved question.

We report here the comprehensive analysis of mice heterozygous for discrete Trio variants associated with different NDDs ASD-associated +/K1431M in the Trio GEF1 domain, SCZ-associated +/K1918X, leading to nonsense decay, and BPD-associated +/M2145T in Trio GEF2 domain. We show that distinct Trio NDD-associated variants differentially impair mouse behavior, brain development, and inhibitory and excitatory synaptic transmission. We show, for the first time, that Trio variants that impact GEF1 and GEF2 activity differentially impact presynaptic neurotransmitter release and synaptic vesicle replenishment. While the Trio K1431M GEF1 domain is impaired in Rac1 activation, unexpectedly +/K1431M mice exhibit increased Rac1 activity, associated with increased levels of the Rac1 GEF Tiam1. Acute Rac1 inhibition with NSC23766 rescued glutamate release deficits in +/K1431M variant cortex. Together, our data show how discrete heterozygous TRIO variants that differentially impact Trio biochemical activities yield divergent effects on behavior, neurodevelopment, and synaptic transmission.

Materials and methods

Animal work

Animal work was compliant with federal guidelines and approved by the Yale Institutional Animal Care and Use Committee. Age-matched male and female mice were used for behavioral experiments; males were used for electrophysiological and neuroanatomical studies to reduce potential variation due to estrus cycle (34).

Generation of Trio mutant mice and mouse genotyping

Mice heterozygous for Trio variants K1431M, K1918X, or M2145T were generated via CRISPR/Cas-mediated genome editing and maintained on a C57Bl/6 background. Details on allele design and genotyping are included in Supp. Methods.

In vitro GEF Assays

The K1431M point mutant was generated via site-directed mutagenesis of human WT TRIO GEF1 using the following oligos: 5’-cagcgaataacgatgtatcagctcc-3’ and 5’-ggagctgatacatcgttattcgctg-3’. Recombinant WT and K1431M human TRIO GEF1 and Rac1 were purified from bacteria as described (35). GEF activity was monitored by the decrease in fluorescent signal (λexcitation = 488 nm; λemission = 535 nm) as GTP was exchanged for BODIPY-FL-GDP on Rac1 over 30 min, as described (33, 35).

Body weight, brain weight, and head width measurements

Ear-to-ear head width of P42 mice was measured with calipers in anesthetized mice. Brain weight was measured after removing the olfactory bulbs and brain stem.

Mouse brain lysates and cortical synaptosome preparation

Whole brain lysates from P0-P1 or P35-P42 mice, and cortical synaptosomes from P39-P42 mice, were prepared as previously described (19, 36, 37), with minor modifications (described in Supp. Methods). Protein concentrations were determined with BCA assay (Pierce).

Western blot and quantification

Brain lysates were separated on SDS-PAGE gels, blotted to nitrocellulose membranes, stained with Ponceau S, blocked in 5% nonfat milk in TBS-T, incubated with primary antibodies overnight at 4°C, then with conjugated secondary antibodies at RT for 1 h (see Supp. Table 1 for list of antibodies). Images captured by a ChemiDoc Imaging System (Bio-Rad) were quantified in ImageJ. Signal intensity was normalized to Ponceau S, then to the WT average.

G-LISA

GTP-bound active Rac1 and RhoA levels in were measured using G-LISA kits according to manufacturer’s instructions (Cytoskeleton, Inc., BK128; Inc., BK124).

Behavioral tests

Behavioral tests were performed in mice 6-8 weeks of age (P42-P56), as previously described (19).

Animal perfusion and tissue processing

Mice were transcardially perfused with heparinized PBS followed by 4% paraformaldehyde (PFA) in phosphate buffered saline (PBS). Brains were postfixed in 4% PFA at 4°C for 24 h, then sliced coronally at 30 μm for Nissl stain and immunohistochemistry and 200 μm for dendritic arbor and spine analysis. For details on tissue processing, see Supp. Methods.

Electron microscopy

Electron microscopy of synapses in M1 layer 5 was performed as previously described (19) and synaptic features were quantified in ImageJ as previously described (38–41) from 5-18 fields of view (∼50 μm2) per mouse by a reviewer blinded to genotype.

Electrophysiological recordings

Acute slices were prepared from mice at P35-42 as previously described (19) with modifications noted here. Coronal slices of M1-M2 cortex were cut at 360 µm in ice-cold N-Methyl-D-glucamine-aCSF (NMDG-aCSF, in mM): 120 NMDG, 2.5 KCl, 7 MgSO4, 1.25 NaH2PO4, 0.5 CaCl2, 28 NaHCO3, 13 glucose, 7 sucrose, saturated with 95% O2-5% O2 at 300-320 mOsmol/L, pH 7.4) on a vibratome (Leica) and recovered in normal ACSF at 32°C for 30 min followed by 1 h recovery at RT. Recorded signals were acquired at 100kHz sampling rate and low pass filtered at 6 kHz. Cells were excluded if series resistance changed >20%.

AMPAR- and NMDAR-eEPSCs, miniature EPSCs (mEPSCs), Long term-potentiation (LTP) and were recorded and analyzed as previously described in (19). Gamma-aminobutyric acid receptor and glycine receptor miniature inhibitory postsynaptic currents (GABAR/GlyR-mIPSCs) were recorded at +15 mV.

Paired-pulse stimulation was applied and paired-pulse ratio (PPR) was calculated as described in (19). Calculations of the readily releasable pool (RRP) size, probability of release (Pr) and synaptic vesicles (SVs) depletion and RRP recovery (fractional) rate were made from recordings of AMPAR-eEPSCs under high frequency stimulation (HFS, 40Hz; 15 pulses) for each neuron as previously described (42–44) (details see in Supp. Methods).

Rescue experiments were performed as described above with addition of 100 µM NSC23766 into the recording solution with 5 min incubation period to allow for efficient slice penetration. All train data were processed and analyzed using OriginLab 10 software.

Quantitative Proteomics

Tissue from P21 mouse cortex (4 per genotype) was processed for tandem mass tag (TMT) peptide labeling as previously described (19) with minor modifications. Desalted 16-plex TMT-labeled peptides were separated by HPLC, and proteome analysis was performed using an EASY-nLC 1200 UHPLC coupled to an Orbitrap Eclipse mass spectrometer (Thermo Fisher Scientific). MS2 spectra were searched using Spectrum Mill (Agilent) against the UniProt Mouse Database. Technical details are described in Supp. Methods.

Gene Set Enrichment Analysis

Mouse UniProtIDs for identified proteins were converted to human orthologs and ranked by signed log10(nominal p-value) (with sign indicating direction of fold-change from WT). GSEA 4.3.3 (45, 46) was used to run the GSEAPreranked tests against all gene sets in the Human MSigDB Collection C2 v2023.2. SynGO 1.2 (47) was used to identify synaptic ontologies for differentially expressed proteins with nominal p-value <0.05.

Data Analysis

Statistical analyses were performed using GraphPad Prism 10. Data in bar graphs are presented as mean ±SEM, with individual data points graphed when applicable. Sample size ‘n’ is annotated within the bars or in the figure legend for each group. Distributions were tested for normality and outliers were removed before proceeding with statistical tests. Specific details of statistical tests performed, adjusted with post-hoc Bonferroni test for multiple comparisons (MC) where appropriate, are included in the figure legends. Significance was defined by a p-value less than 0.05: nsp<0.1; *p<0.05; **p<0.01; ***p<0.001; ****p<0.0001.

Results

Generation of Trio variant mice and impact on Trio isoforms

We used CRISPR/Cas9 technology to generate mice bearing heterozygous Trio rare variant alleles associated with ASD (+/K1431M) and SCZ (+/K1918X), and a de novo mutation found in an individual with BPD (+/M2145T) (Fig. 1A,B). We chose these alleles for their discrete effects on TRIO levels or activity: K1431M impairs TRIO GEF1 activity in vitro up to 8-fold (Supp. Fig. 1A,B) (24), M2145T TRIO GEF2 has reduced ability to activate RhoA as a function of protein concentration in cells (24), and K1918X creates a premature stop codon between GEF1 and GEF2.

Genetically engineered mice with heterozygosity for K1431M, K1918X, or M2145T Trio variants have divergent effects on Trio protein expression, Rho GTPase activity, and NDD-like behaviors.

(A) Schematic of major Trio isoforms present in the adult mouse brain, with locations of engineered neurodevelopmental disease (NDD)-associated Trio variants: K1431M autism spectrum disorder (ASD)-associated missense variant in the GEF1 DH domain; K1918X schizophrenia (SCZ)-associated nonsense variant that lies just before the GEF2 domain; and M2145T bipolar disorder (BPD)-associated missense variant in the GEF2 DH domain. (B) Representative sequencing chromatograms of wild-type (WT), +/K1431M, +/K1918X, and +/M2145T mice. Arrows indicate heterozygosity for the variant alleles. (C) Representative immunoblots for Trio in P0 brain lysates using an antibody against Trio spectrin repeats (SR5-6). (D) Quantification of Trio protein levels from P0 brain lysates. Trio protein levels are reduced only in the brains of +/K1918X mice compared to WT controls (0.545 ±0.126 of WT level, p=0.0046). (E-H) Activity levels of Rac1 (E,G) and RhoA (F,H) in whole brain homogenates of neonate (P0, E-F) and adult (P42, G-H) Trio variant mice as measured by G-LISA assay. Rac1 activity is increased in +/K1431M mice relative to WT at both ages (1.106 ±0.027-fold at P0, p=0.0035; 1.509 ±0.175-fold at P42, p=0.0279) and decreased in neonate +/K1918X mice (0.908 ±0.0.032-fold, p=0.0230), with a trend towards increased activity in adult +/M2145T mice (1.438 ±0.183-fold, p=0.0843); meanwhile RhoA activity appears unchanged in all mice relative to WT, though there may be a trend towards decreased activity in +/K1918X neonates (0.840 ±0.074-fold, p=0.1292). (I,J) Activity levels of Rac1 (I) and RhoA (J) in synaptosomes isolated from P42 mouse cortex. Rac1 activity is increased in +/K1431M synaptosomes (1.125 ±0.107-fold, p=0.0023), while RhoA activity is decreased in +/M2145T synaptosomes (0.731 ±0.042-fold, p=0.0093) relative to WT. Data are presented as mean ±SEM. For (D-J), one-way ANOVA with post-hoc Bonferroni MC test identified differences from WT (nsp<0.1, *p<0.05, **p<0.01). Numbers of mice quantified per group are annotated inside the bar. (K) +/K1431M and +/K1918X male mice had decreased latency to fall off an accelerating rotarod compared to WT male mice. Two-way ANOVA with post-hoc Bonferroni MC test identified differences from WT for each individual trial (*p<0.05, **p<0.01, ***p<0.001, ****p<0.0001). Linear regressions identified differences from WT in slopes (WT 16.96 ±1.344; +/K1431M 7.270 ±2.019, p<0.0001; +/K1918X 10.61 ±1.444, p<0.0001; ####p<0.0001). n=40 WT; 10 +/K1431M; 16 +/K1918X; 13 +/M2145T male mice. (L) WT mice displayed normal social preference and spent more time with a stranger mouse (Str.) relative to an inanimate object (Obj.) in a three-chamber interaction test. Social interactions were impaired in +/K1431M male mice. (M) WT mice spent more time with a novel object (N) than with a familiar object (F). Novel object recognition was impaired in +/K1918X male mice. All data are presented as mean ±SEM. For (L-M), two-way ANOVA with post-hoc Bonferroni MC test identified differences from WT (nsp<0.1, *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001). Numbers of mice quantified per group are annotated inside the bar. (N) Male +/K1918X mice exhibited increased nestlet shredding over 30 min (26.26 ±3.61% shredded vs WT 14.26 ±2.97%; p=0.0433), and +/K1431M mice exhibited a trend toward increased nestlet shredding (25.90 ±4.34% shredded, p=0.1038) compared to WT mice. One-way ANOVA with post-hoc Bonferroni MC test identified differences from WT (nsp<0.1, *p<0.05). n=19 WT; 10 +/K1431M; 15]+/K1918X; 9 +/M2145T male mice.

Mice heterozygous for these Trio alleles survived to adulthood. Mice homozygous for Trio K1431M and M2145T survived to adulthood, with genotypes from offspring of +/K1431M intercrosses observed in Mendelian ratios, but fewer than expected M2145T homozygotes obtained (Supp. Fig. 1C). K1918X homozygote pups were not observed, as expected for a null allele (16, 18). We focused on heterozygotes, as most rare damaging Trio variants are heterozygous in humans. Heterozygosity for the K1431M and M2145T alleles did not alter the levels of the predominant Trio isoforms (Trio9S, 263 kDa; Trio9L, 277 kDa; Duet, 145 kDa; cerebellum-specific Trio8 217 kDa (48, 49)) in the brain at postnatal day 0 (P0) (Fig. 1C-D) or P42 (Fig. 1A, Supp. Fig. 1D-L). Trio9 protein levels were reduced by ∼50% in the brains of +/K1918X mice at P0 (Fig. 1C-D) and P42, and we did not detect residual truncated protein (expected 217 kDa), suggesting this mutation leads to nonsense-mediated decay (Supp. Fig. 1D-L); levels of Trio8 and Duet were unaffected (Supp. Fig. 1D-L).

Trio variant alleles differentially impact active Rho family GTPase levels

Given the effects of these variants on TRIO GEF1/2 activities, we measured levels of active GTP-bound Rac1 and RhoA in brains of neonatal (P0) and adult (P42) Trio variant-bearing mice (Fig. 1E-H). In concordance with the ∼50% reduction in Trio levels in +/K1918X brains, Rac1 activity was decreased in P0 +/K1918X brains (91% of WT activity), with a trend toward decreased active RhoA levels (84% of WT, p=0.0865). By P42, +/K1918X brains did not differ from WT in Rac1 or RhoA activity.

As K1431M decreases TRIO GEF1 catalytic activity in vitro (Supp. Fig. 1A,B) (20, 24, 29), we anticipated Rac1 activity might be reduced in +/K1431M mice. Instead, active Rac1 levels were increased in +/K1431M whole brain lysates compared to WT controls at P0 ((Fig. 1E),111% of WT) and even higher at P42 ((Fig. 1G), 150% of WT). We also did not detect changes in active RhoA levels at the whole brain lysate level in +/M2145T mice at either age (Fig. 1F,H).

Because Trio, Rac1, and RhoA are enriched at synapses (15, 22, 24, 50), we also measured their activities in synaptosomes (Supp. Fig. 1M-P). Rac1 activity was significantly increased in +/K1431M crude synaptosomes from P42 cortex (112% of WT) (Fig. 1I), consistent with measurements in whole brain lysates. Consistent with its reduced GEF2 activity, RhoA activity was decreased in +/M2145T synaptosomes compared to WT (73% of WT) (Fig. 1J).

Distinct Trio variants differentially impact mouse behavior

NDDs affect learning and memory, compulsivity, motor coordination, and social skills, hence we assessed these skills using a diverse array of established behavioral tests (19) in mice bearing Trio variants. +/K1431M and +/K1918X mice of both sexes fell from an accelerating rotarod with reduced latency and improved more slowly in this skill over repeated trials relative to WT littermates (Fig. 1K, Supp. Fig. 1Q), while +/M2145T mice performed similarly to WT. No deficits in muscle strength were noted in any genotype using the Kondziela inverted screen test prior to rotarod testing, and no differences in locomotor or exploratory activity were seen in an open field test (data not shown). In a three-chamber social interaction test, +/K1431M mice of both sexes showed no preference for the stranger mouse over the object (Fig. 1L, Supp. Fig. 1R). +/M2145T mice of both sexes and +/K1918X males exhibited preference for the stranger mouse, similar to WT. In a novel object recognition test, +/K1918X mice of both sexes and +/M2145T females failed to discriminate between the novel and familiar objects, while +/K1431M mice and +/M2145T males exhibited normal discrimination between novel and familiar objects similar to WT (Fig. 1M, Supp. Fig. 1S). +/K1918X males exhibited significantly more compulsive nestlet shredding relative to WT, while +/K1431M males showed a trend toward increased nestlet shredding (Fig. 1N).

Overall, behavioral phenotypes overlapped between +/K1431M and +/K1918X mice, though there were distinct differences in social interaction and memory tests. +/M2145T mice exhibited the fewest phenotypes in measured tasks. Together, these data indicate that these Trio alleles differentially impact behavior.

Trio +/K1431M and +/K1918X mice have smaller brains

TRIO variants that reduce TRIO GEF1 activity are associated with microcephaly (25, 26, 33), so we assessed head and brain size in Trio variant mice (Fig. 2A-E, Supp. Fig. 2A-E). After adjusting for body weight, both the head width and brain weight (Fig. 2D,E) were reduced in adult +/K1431M mice relative to WT, consistent with the microcephaly seen in patients harboring TRIO GEF1-deficient alleles. Head width in +/K1918X mice was reduced, but in proportion to their body weight (head width-to-body weight ratios did not differ between +/K1918X and WT mice, p>0.95). However, +/K1918X mice still displayed a reduced brain weight relative to WT, suggesting a disproportionate reduction in brain mass relative to their already smaller head size. Meanwhile, +/M2145T mice did not differ from WT in either head width or brain weight when adjusted for body weight.

Trio +/K1431M and +/K1918X mice have smaller brain weights, but only +/K1918X brains have smaller less complex neurons.

(A) Ear-to-ear head width is reduced in P42 +/K1918X and +/M2145T compared to WT male mice (+/K1918X 12.125 ±0.074 mm, p=0.0096; +/M2145T 12.012 ±0.152 mm, p=0.0007; vs WT 12.414 ±0.049 mm). (B) Brain weight is significantly reduced in P42 +/K1431M, +/K1918X, and +/M2145T males compared to WT (+/K1431M 0.379 ±0.004 g, p=0.0006; +/K1918X 0.345 ±0.004 g, p<0.0001; +/M2145T 0.381 ±0.005 g, p=0.0016; vs WT 0.399 ±0.004 g). (C) Body weight is increased in P42 +/K1431M and decreased in +/K1918X males (+/K1431M 22.911 ±0.382 g, p=0.0034; +/K1918X 20.700 ±0.338 g, p=0.0153; vs WT 21.745 ±0.224 g). (D) Head widths normalized to body weight of P42 +/K1431M male mice were reduced 11.8% compared to WT mice (0.520 ±0.008 mm/g vs WT 0.589 ±0.011 mm/g, p=0.0001). Head width-to-body weight ratios were calculated per individual mouse, with mouse number per group annotated within the bar. (E) Brain weights normalized to body weight of P42 +/K1431M and +/K1918X male mice were reduced 9.9% and 9.2%, resp., compared to WT mice (+/K1431M 0.0165 ±0.0004, p=0.0002; +/K1918X 0.0167 ±0.0004, p=0.0015; vs WT 0.0184 ±0.0004). Ratios were calculated by dividing the mean brain weight (B) by the mean body weight (C) of overlapping but non-identical populations of mice (see (B) and (C) for mouse number per group). For (A-E), data are presented as mean ±SEM; only male mice are shown here (data for females in Supp. Fig. 2). One-way ANOVA with post-hoc Bonferroni MC test identified differences from WT (nsp<0.1, *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001; n=number of mice per group, annotated within the bar). (F) Representative Nissl-stained image of a 30 μm coronal slice from a P42 WT brain. Magnification of the cortex (dotted black box) reveals cortical layers. (G) Total cross-sectional brain area of Nissl-stained coronal sections was reduced by ∼9% in P42 +/K1918X brains (43.21 ±0.577 mm2 vs WT 47.29 ±0.823 mm2, p=0.0045). (H) Total thickness of the cortex, in the region marked by the dotted black box in (F), is reduced by ∼8% in P42 +/K1918X brains (1.304 ±0.02262 mm vs WT 1.416 ±0.01417 mm, p=0.0021). For (G,H), data are presented as mean ±SEM. Ordinary one-way ANOVA with post-hoc Bonferroni MC test identified differences from WT (**p<0.01; n=number of mouse brains, annotated within the bar). (I) Representative maximum projection fluorescence image and corresponding dendritic arbor reconstruction of a motor cortex Layer 5 pyramidal neuron (M1 L5 PN) from a WT mouse. (J,K) Dendritic field size as measured by convex hull analysis of basal (J) and apical (K) dendrite arbor reconstructions of M1 L5 PNs of P42 Trio variant mice. +/K1918X trended toward smaller basal dendritic field size (0.1172 ±0.0078 mm2 vs WT 0.1368 ±0.0077 mm2, p=0.0933), and both +/K1918X and +/M2145T had significantly smaller apical dendritic field size (+/K1918X 0.5157 ±0.0169 mm2, p=0.0460; +/M2145T 0.4893 ±0.0285 mm2, p=0.0062; vs WT 0.6081 ±0.0319 mm2) compared to WT L5 PNsOrdinary one-way ANOVA with post-hoc Bonferroni MC test (B) identified differences from WT (nsp<0.1, *p<0.05, **p<0.01). (L-M) Sholl analysis revealed basal (L) and apical (M) dendritic arborization changes in Trio variant M1 L5 PNs compared to WT: both basal and apical arborization was reduced in +/K1918X, while proximal basal arborization was increased in +/K1431M. Two-way ANOVA (stacked) with post-hoc Bonferroni MC test identified differences from WT at each radius centered at the soma (*p<0.05, **p<0.01, ***p<0.001) For (J-M), n=23 neurons from 17 WT mice, 22 neurons from 15 +/K1431M mice, 20 neurons from 12 +/K1918X mice, 20 neurons from 14 +/M2145T mice.

Trio variant mice show limited changes in cortical organization

We examined whether the decreased brain weight and head size observed in +/K1431M and +/K1918X mice were associated with anatomical defects. Histological analyses did not reveal gross morphological defects (data not shown). We observed reductions in both total cross-sectional brain area and cortical thickness in +/K1918X brains (Fig. 2F-H), consistent with their smaller head size and brain weight. Significant decreases in cortical layer 2/3 and layer 5 thickness were observed in +/K1918X brains; but these were in proportion to the relative decrease in cortical thickness as compared to WT (Supp. Fig. 2F-H). No changes were observed in +/M2145T mice.

No differences in total cortical cell density or in layer-specific cell density were observed in Trio variant mice relative to WT, nor were numbers or proportions of cortical NeuN+ neuronal cells or PV+ inhibitory neurons altered relative to WT, although there were trends toward increased cell density in +/K1918X cortex and increased NeuN+ cell density in +/M2145T motor cortex (Supp. Fig. 2M-Q). These data suggested that the reduced brain size +/K1918X mice resulted from a loss of neuropil.

Trio variant heterozygotes exhibit alterations in dendritic arbors and synaptic ultrastructure

Altered dendritic arbor morphology and dendritic spine abnormalities are hallmarks of NDDs (51–57). Excitatory neuron-specific ablation of one or both Trio alleles decreased dendritic arborization, increased spine density, and yielded smaller synapses in cortex area M1 Layer 5 pyramidal neurons (M1 L5 PNs) (19).

Sholl analysis of M1 L5 PNs revealed significant reductions in both basal and apical dendritic arbor complexity and dendritic field size in P42 +/K1918X neurons, an increase in proximal basal arbor complexity in +/K1431M neurons, and a slight decrease in distal apical arbor complexity in +/M2145T neurons relative to WT (Fig. 2I-M). Path length analysis revealed that branch numbers were increased for lower order dendrites in +/K1431M neurons, while average branch lengths were decreased at higher order dendrites in +/K1918X neurons (Supp. Fig. 3F-K). None of the Trio variant heterozygotes exhibited altered dendritic spine density on M1 L5 pyramidal neurons compared to WT mice on either apical or basal arbors (Supp. Fig. 3L,M).

Electron microscopy of cortical area M1 L5 revealed that synapse density was significantly increased in +/K1918X mice compared to WT (Fig. 3A,B), possibly due to a net reduction in neuropil resulting from smaller dendritic arbors. Within synapses, postsynaptic density (PSD) length was slightly decreased in +/K1918X and +/M2145T mice, but not in +/K1431M mice (Fig. 3C). Cross-sectional presynaptic bouton area and spine head area were similar to WT in each Trio variant heterozygote (Fig. 3D,E).

Trio variants differentially impact synapse ultrastructure and synaptic vesicle distribution.

(A) Representative electron micrograph (EM) images from motor cortex layer 5 (M1 L5) of P42 WT and Trio variant mice. Post-synaptic regions are pseudo-colored in cyan; pre-synaptic regions in magenta. Scale bar = 0.5 μm. (B) Asymmetric synapses from all EM images of M1 L5 were quantified (from 18 fields of view per mouse, 5 mice per group). Synapse density was increased in +/K1918X mice (0.09205 ±0.004775 synapses/um2; vs WT 0.07633 ±0.003954 synapses/um2, p=0.0345). (C) PSD lengths were decreased in M1 L5 synapses of +/K1918X and +/M2145T mice (+/K1918X 0.2926 ±0.004652 um, p=0.0204; +/M2145T 0.2916 ±0.004922 um, p=0.0142; vs WT 0.3125 ±0.005612 um). (D,E) Presynaptic bouton and spine head areas of M1 L5 synapses were unchanged from WT in all Trio variants. For (C-E), number of measurements per mouse across all 5 mice per genotype are denoted in the corresponding figure. (F) Synaptic vesicles (SVs) counted per 100 nm of active zone (AZ) length in M1 L5 as a function of distance from the AZ. +/M2145T showed an increase in the readily releasable pool (RRP) identified as docked SVs (15 nm from AZ; 1.23 ±0.05 vs. WT 0.90 ±0.05) and an increase in tethered SVs (50 nm from AZ; 1.44 ±0.04 vs. WT 1.20 ± 0.05). +/K1918X and +/M2145T also showed an increase in the reserve pool of SVs (200 nm from AZ; 3.51 ±0.21 and 3.81 ±0.18, resp. vs WT 2.74 ±0.16). (G) Total releasable pool, calculated as the number of SVs at 15-150 nm from AZ per area of distribution (nm2). RRP was significantly increased in +/M2145T (0.257 ±0.007 vs WT 0.228 ±0.008), driven by increased docked and tethered SVs. All data are presented as mean ±SEM. Ordinary one-way ANOVA with post-hoc Bonferroni MC test identified differences from WT (*p<0.05, **p<0.01, ***p<0.001, ****p<0.0001). For (F,G), 35-50 synapses per mouse were analyzed.

Synaptic vesicle (SV) distribution was significantly altered in +/M2145T mice relative to the other genotypes (Fig. 3F,G). Within the total releasable pool 15-150 nm from the active zone (AZ), both docked SVs at 15nm comprising the readily-releasable pool (RRP) (40, 41) and tethered SVs at 50 nm were significantly increased in density in +/M2145T mice relative to the other genotypes (Fig. 3F,G). SVs distribution at 200 nm from the AZ which contribute to the reserve pool were also significantly increased in +/M2145T, suggesting an overall SV pool size. No differences in synaptic vesicle distribution were noted in +/K1431M mice, while +/K1918X showed modestly increased SVs at 200 nm from AZ relative to WT.

Synaptic transmission and plasticity are differentially impaired by distinct Trio variants

Loss of Trio function or disruption of TRIO GEF1 activity in slice culture decreases AMPAR levels at excitatory synapses (19, 21, 22, 29), while mice bearing a GEF1-deficient Trio allele exhibited decreased gamma-aminobutyric acid receptor (GABAR)- and glycine receptor (GlyR)-mediated inhibitory miniature current (mIPSC) frequencies in the prefrontal cortex (20). To explore how Trio variants impact synaptic function, we measured both miniature excitatory currents (mEPSCs) and mIPSCs in M1 L5 PNs in each of the Trio variant heterozygotes.

AMPAR-mediated mEPSC amplitudes were significantly increased in +/K1431M and +/K1918X mice relative to WT littermates, with no change in their frequencies (Fig. 4A-C). In contrast, AMPAR-mediated mEPSC amplitudes were unchanged in +/M2145T mice, but their frequencies were significantly increased. No significant changes in NMDAR mEPSC amplitudes were noted, while NMDAR mEPSCs frequencies were decreased in +/K1431M and increased in +/M2145T mice cortex (Fig. 4D-F). Notably, these findings correlated with gross alterations in the ratio of NMDAR/AMPAR-mediated evoked (e)EPSCs measured in M1 L5 PNs following stimulation in L2/3. All Trio variant heterozygotes showed decreased NMDA/AMPA ratios, indicating imbalances in NMDAR-versus AMPAR-mediated conductance (Supp. Fig. 4A,B). Significant decreases in mIPSC frequencies were noted in +/K1431M and +/M2145T mice relative to WT mice, with no change in amplitudes (Fig. 4G-I). +/K1918X mice exhibited increased mIPSC amplitude with no observed change in mIPSC frequency. Together, these data indicate that the Trio variants differentially impact excitatory and inhibitory transmission.

Trio variant mice exhibit deficits in synaptic signaling and LTP.

(A,D) Representative traces of miniature excitatory AMPAR-, NMDAR-mediated mEPSCs and (G) inhibitory postsynaptic currents mIPSCs in M1 L5 pyramidal neurons of WT, +/K1431M, +/K1918X and +/M2145T mice. (B) AMPAR-mediated mEPSC amplitudes (I) were significantly increased in +/K1431M (16.67 ± 1.04 pA) and +/K1918X (14.71 ± 0.92 pA) slices, with no observed changes in +/M2145T slices (13.90 ± 1.16 pA) compared to WT (11.25 ± 0.84 pA; n=17-25 neurons from ≥6-8 mice per group). (C) No significant changes in AMPAR mEPSC frequencies (q) were observed in +/K1431M and +/K1918X, while +/M2145T had an increase (2.20 ± 0.15 1/s; vs WT 1.55 ± 0.09 1/s). (E, F) NMDAR mEPSC frequencies were reduced in +/K1431M (0.89 ± 0.12 1/s; vs WT 1.3324 ± 0.11 1/s; n=9-13 neurons from ≥ 5-7 mice per group) and showed a slight but significant increase in +/M2145T mice (1.68 ± 0.10 1/s). (H, I) GABA/GlyR mIPSC amplitudes were significantly increased in +/K1918X vs WT (23.69 ± 2.89 pA; vs 15.86 ± 1.56 pA, resp.), while frequencies were decreased in +/K1431M and +/M2145T (0.94 ± 0.14 1/s and 1.64 ± 0.19 1/s resp.; vs WT 2.44 ± 0.20; n=16-26 neurons from ≥6-8 mice per group). For (B-I), data are presented as mean ± SEM. One-way ANOVA with post-hoc Bonferroni test identified differences from WT (*p<0.05; **p<0.01; ***p<0.001; ****p<0.0001). (J) Averaged representative traces of baseline and post-TBS eEPSCs currents in M1 L5 PNs of WT and Trio variant mice. (K) Normalized eEPSCs amplitudes measuring LTP in L5 PNs by TBS in L2/3 afferents in all genotypes showed a significant decrease in the initiation and no potentiation of the LTP in +/K1431M and +/K1918X, with the increase in initiation and potentiation of +/M2145T M1 L5 PNs compared to WT. LTP was induced at 0 min. RM two-way ANOVA with post-hoc Bonferroni MC test identified differences from WT (****p<0.0001; n=6-8 neurons from ≥4-5 mice per group).

Finally, we tested the ability of the L5 PNs to undergo long-term potentiation (LTP) following theta-burst stimulation of L2/3 afferents (Fig. 4J,K). While LTP was robustly induced and potentiated in M1 L5 PNs from WT mice, LTP induction and potentiation were deficient in slices from +/K1918X and +/K1431M mutant mice. In contrast, +/M2145T L5 PNs showed increased induction and prolonged potentiation of LTP compared to WT L5 PNs.

Excitatory neurotransmitter release is altered in Trio +/K1431M and +/M2145T heterozygotes

In addition to its postsynaptic roles, Trio localizes presynaptically and interacts with the presynaptic active zone scaffolding proteins Bassoon and Piccolo (14, 15, 22). To assess the impact of Trio variants on presynaptic function, we first measured the paired-pulse ratio (PPR). Synapses with low probability of release (Pr), e.g. L2/3 synapses onto L5 PNs, can undergo paired-pulse facilitation (PPF), a relative increase of synaptic transmission in response to the second of two closely-apposed stimuli. WT and +/K1918X M1 L5 PNs exhibited normal facilitation of eEPSC amplitudes that decreased with increased interstimulus interval (ISI) (Fig. 5A,B). PPF was significantly enhanced in +/K1431M M1 L5 PNs at short ISIs relative to WT, suggesting a possible reduction in glutamate Pr (Fig. 5A,B). The +/M2145T PPR curve was unusual, with significantly reduced PPF at short ISIs, yet clearly increased PPF at longer ISI (Fig. 5A,B) compared to WT. The decreased PPF at initial ISI in +/M2145T mice correlated with increased mEPSC frequency (Fig. 4A-C), suggestive of a possible increase in spontaneous glutamate Pr.

Trio GEF damaging variant mice have deficiencies in excitatory neurotransmitter release.

(A) Representative traces recorded from M1 L5 PNs of WT, Trio variant mice, and rescue experiment in acute slices of +/K1431M with 100 µM NSC23766 (NSC, Rac1 inhibitor) in response to paired-pulse stimulation in M1 L2/3. (B) Paired-pulse ratio (PPR) at varying interstimulus intervals (ISI), overlaid with a single exponential fit (except for +/M2145T data). An increase in the initial PPR was observed in M1 L5 PNs of +/K1431M mice (35 ms: 1.70 ± 0.089, p=0.003; 60 ms: 1.40 ± 0.07, p=0.046; 100 ms: 1.27 ± 0.05, p=0.031; n=20-34 neurons from ≥7-9 mice per group) with no changes in +/K1918X. In +/M2145T slices, we observed a decrease in initial PPR at shorter ISI (35 ms: 1.05 ± 0.06, p<0.0001; 60 ms: 0.97 ± 0.06, p=0.037) and an increase at longer ISI (100 ms: 1.36 ± 0.09, p=0.034; 200 ms: 1.18 ± 0.08, p=0.013) compared to WT (35 ms: 1.40 ± 0.04; 60ms: 1.21 ± 0.03; 100 ms: 1.13 ± 0.03; 200 ms 1.0 ± 0.02; 300 ms 0.96 ± 0.17). Acute application of NSC onto +/K1431M slices significantly shifted the PPR curve downwards at all points compared to untreated +/K1431M slices and showed no significant difference from WT (+/K1431M + NSC 35 ms: 1.25 ± 0.06, p<0.0001; 60 ms: 1.13 ± 0.052, p=0.0007; 100 ms: 1.02 ± 0.053, p=0.0017; 200 ms 0.91 ± 0.039, p=0.0043; 300 ms 0.88 ± 0.045, p=0.021). (C) Representative traces of AMPAR eEPSCs in M1 L5 PNs under HFS (15 pulses at 40 Hz) in L2/3. (D) AMPAR eEPSCn amplitudes normalized to eEPSC1 of the train revealed changes in the depletion rates during HFS in Trio variants compared to WT (tau decay (τd), WT: 2.70 s, +/K1431M: 3.19 s, +/M2145T: 4.79 s, +/K1918X: 3.52 s, +/K1431M + NSC: 2.68 s; n=12-15 neurons from 5-7 mice). (E) The estimated glutamate probability of release (Pr) was decreased in +/K1431M (0.13 ± 0.099) and increased in +/M2145T (0.26 ± 0.019), with no significant change in +/K1918X (0.15 ± 0.01) compared to WT M1 L5 PNs (0.19 ± 0.01; n=11-15 neurons from ≥ 5 mice per group); acute NSC application rescued Pr in +/K1431M (0.22 ±0.019; n=11-15 neurons from ≥ 5 mice per group). (F) The calculated size of the readily releasable vesicle pool (RRP) was increased in +/M2145T M1 L5 PNs compared to WT (665.7 ± 68.5 pA vs 415.8 ± 43.9 pA). RRP in +/K1431M synapses before or after treatment with NSC did not differ from WT (543.1± 64.4 pA; + NSC: 427.9± 79.2 pA vs 415.8 ± 43.9 pA) (G) Exponential fits of the fractional recovery plotted vs ISI. Time of recovery, measured by exponential tau recovery (τR), was significantly decreased in +/K1431M M1 L5 PNs (6.3 s, vs WT 1.7 s). +/K1431M exhibited an inability to fully recover to initial levels after ISI 18 s, while NSC application has allowed for full recovery at 18s. It improved but did not fully rescue tau recovery time in +/K1431M (5.2 s). Data are presented as mean ± SEM, with significant differences from WT tested using one-way ANOVA with post-hoc Bonferroni (*p<0.05; **p<0.01; ***p<0.001; ****p<0.0001).

We used high frequency (40Hz) stimulation (HFS) trains to quantitatively estimate glutamate Pr, RRP size, and rates of SV depletion and recovery. A plot of the normalized eEPSC responses to a HFS train stimulation again revealed facilitation upon the first 2-3 stimulations that was increased in +/K1431M and decreased in +/M2145T slices relative to WT, with no changes in +/K1918X slices (Fig. 5C,D). Initial facilitation was followed by decaying eEPSC amplitudes, reflecting SV depletion under HFS. All Trio variants exhibited a slower train decay rate relative to WT during HFS with +/M2145T depleting at half the rate of WT (τd, 4.8 s vs WT: 2.7 s) (Fig. 5D).

We used a ‘Decay’ method (42, 43) to estimate Pr and RRP size from HFS trains. Glutamate Pr in the L2/3-L5 synapses was increased in +/M2145T mice, while it was decreased for +/K1431M mice (Fig. 5E), consistent with the relative changes observed in PPF for these mice. RRP size was much larger in L2/3-L5 synapses of +/M2145T mice relative to WT (Fig. 5F), consistent with the increased SV distribution found in close apposition to the AZ (Fig. 3F,G). We tested the ability of the Trio variant heterozygotes to recover after train depletion by pairing HFS trains with a single stimulus at increasing intervals (0.01, 2, 6, 9, 12, 18s). The recovery rate (τR) was significantly slower in +/K1431M L5 PNs and they did not recover to their initial strength within 18 s, plateauing at 78% of maximal recovery compared to WT (Fig. 5G). Altogether, +/M2145T and +/K1431M L2/3-L5 synapses show a deficiency in both synchronous and spontaneous glutamate release as identified by HFS and frequency changes of mEPSCs.

Trio variant cortex display different proteomic signatures

We used comparative proteomics from P21 cortex to identify proteins and pathways that were differentially altered by the Trio variants. We quantified a total of 7,362 proteins, finding distinct differences in the cortical proteome for each genotype (Supp. Fig. 5, Supp. Table 2). Gene Set Enrichment Analysis (GSEA) (45, 46) revealed alterations in distinct functions for each Trio variant (Fig. 6A). Of note, the only enriched gene set specific to neurons was downregulation of the synaptic vesicle pathway in +/K1431M cortex; all other gene sets were not cell type-specific.

Trio variant mice show different molecular changes in the cortex involving presynaptic machinery and Rac1 GEFs.

(A) Bar graph illustrating the top enriched pathways (FDR q-value<0.2, *FDR<0.05) identified by gene set enrichment analysis (GSEA) for each Trio mutant mouse compared to WT, using all proteins quantified by mass spectrometry in P21 cortex (7362 proteins, n=4 mice per genotype), sorted by normalized enrichment score (NES). Pathways with a positive NES are upregulated compared to WT; pathways with a negative NES are downregulated compared to WT. [] denotes the gene set: [R] = Reactome, [WP] = WikiPathways, [K] = KEGG. (B,C) Bar graphs illustrating the top enriched (FDR q-value <0.001), (B) cellular components, and (C) biological processes identified by GSEA, using synaptic proteins from SynGO gene sets (n=1077 proteins). For a complete list of all enriched pathways identified by GSEA and SynGO, see Supp. Table 3. (D) Representative immunoblots for Munc18-1, synaptophysin (Syp), syntaxin-1a (Stx1a), and synaptotagmin3 (Syt3) in synaptosomes isolated from P42 cortex of WT and Trio variant mice. (E-H) Normalized intensity levels from immunoblots of select presynaptic proteins from P42 cortical synaptosomes. Munc18-1, Syp, and Syt3 are significantly increased in +/M2145T synaptosomes compared to WT; Syp is increased while Stx1a is significantly decreased in +/K1431M synaptosomes compared to WT. Ordinary one-way ANOVA with post-hoc Bonferonni MC test identified differences from WT (*p<0.05, **p<0.01, ***p<0.001; n=synaptosomes from 14 WT, 10 +/K1431M, 9 +/K1918X, and 7 +/M2145T male mice). (I) Representative immunoblots for Kalirin, Tiam1, and VAV2 in P42 cortical RIPA lysates of WT and Trio variant mice. (J-L) Normalized intensity levels from immunoblots of select RhoGEFs from P42 cortical lysates. Kalirin levels are unchanged in all Trio variant cortex compared to WT; Tiam1 levels are increased ∼47% in +/K1431M and increased ∼45% in +/M2145T cortex compared to WT; VAV2 is increased ∼34% in +/M2145T cortex compared to WT. Unpaired t-tests identified differences from WT (*p<0.05; n=6 mice per genotype).

We used SynGo (47) to investigate whether the Trio variants impacted synaptic functions. 1,067 of the 7,362 total quantified proteins were synaptic proteins listed in the SynGO geneset. When restricted to brain-specific genes, all three Trio variant heterozygotes showed enrichment of differentially expressed proteins (DEPs) in synaptic processes. Notably, +/M2145T upregulated DEPs and +/K1431M downregulated DEPs were significantly enriched at the presynapse, but with only +/K1431M downregulated DEPs showing significant enrichment for postsynaptic neurotransmitter receptor and synaptic vesicle cycling (Fig. 6B,C). Together, our proteomics data point to a significant deficit in presynaptic function in both +/K1431M and +/M2145T cortex, as well as a significant effect of +/K1431M on postsynaptic function.

Rho GEFs and synaptic regulatory proteins are altered in Trio +/K1431M and +/M2145T heterozygotes

Given our findings from proteomic analysis and electrophysiology, we measured levels of key presynaptic regulators, including synaptophysin (Syp), syntaxin binding protein1 (Stxbp1, also known as Munc18-1), syntaxin1a (Stx1a) and synaptotagmin3 (Syt3), which are crucial for synaptic vesicle (SV) tethering, docking, replenishment, and calcium-dependent replenishment, respectively. We also measured levels of presynaptic proteins in P42 cortical synaptosomes, where these proteins are enriched. Munc18-1, Syt3, and Syp levels were increased in +/M2145T synaptosomes relative to WT (Fig. 6D-H). Meanwhile, Stx1a levels were significantly decreased in +/K1431M synaptosomes compared to WT, with no significant changes in +/K1918X compared to WT mice.

The elevated Rac1 activity in +/K1431M brain lysates and synaptosomes (Fig. 1I) seemed at odds with previous reports that K1431M reduces TRIO GEF1 activity (Supp. Fig. 1A,B) (20, 24, 29). We hypothesized that homeostatic compensation in +/K1431M mice may alter expression of other RhoGEFs and GAPs. Indeed, levels of the Rac1 GEF Tiam1 were increased in both +/K1431M and +/M2145T P42 cortical lysates, while VAV2 levels were increased in +/M2145T P42 lysates (Fig. 6I-L). Levels of the Trio paralog Kalirin (58) were unaffected in the Trio variant mice at P42. Together, our proteomic analyses suggest that presynaptic functions are altered in +/K1431M and +/M2145T mice and may be driven by abnormal levels of crucial presynaptic regulatory proteins and changes in Rac1 and RhoA activity.

NSC23766, a Rac1-specific inhibitor, rescues neurotransmitter release in Trio +/K1431M heterozygotes

Rac1 negatively regulates synaptic vesicle replenishment and synaptic strength in excitatory synapses (15, 44). Rac1 activity was increased while Stx1a levels were decreased in +/K1431M cortex, in association with reduced synaptic strength and deficient vesicle replenishment in L2/3-L5 cortical synapses. We tested if the acute application of the Rac1 inhibitor NSC23766 (NSC) could rescue these deficits. Treatment of +/K1431M slices acutely with NSC normalized PPR in M1 L5 PNs at all ISIs to a WT pattern, suggesting a rescue of the decreased Pr (Fig. 5A,B). Application of NSC to +/K1431M slices also decreased time to depletion of signal upon HFS (Fig. 5C,D) and significantly increased Pr without effect on RRP size (Fig. 5E,F), consistent with prior work showing that changes in Rac1 signaling does not impact the RRP (15, 44, 59). Finally, we tested if acute Rac1 inhibition impacts the rate of recovery of the RRP following HFS train stimulation. The recovery curve exhibited improvement at later intervals with full recovery at 18s interval under NSC treatment, but while the recovery rate (τR) measured was improved, it remained significantly slower compared to WT (Fig. 5G). Overall, we demonstrate that presynaptic Trio GEF1-dependent Rac1 signaling is crucial for maintaining synchronous glutamate Pr and SV replenishment at cortical L2/3-L5 synapses.

Discussion

Large-scale genetic studies show significant overlap in risk genes for ASD, SCZ, and BPD, many converging on synaptic proteins (60–68). However, how variants in a single gene contribute to different NDDs remains a major unresolved question. Our study of mice heterozygous for Trio variants associated with ASD (+/K1431M), SCZ (+/K1918X), and BPD (+/M2145T) revealed that the lesions differentially affect Trio protein levels or GEF activity, yielding overlapping but distinct behavioral, neuroanatomical, and synaptic phenotypes. Our findings extend prior work demonstrating that Trio is critical for postsynaptic signaling and synaptic plasticity. We also demonstrate for the first time in mouse model that Trio is critical for glutamate release and synaptic vesicle recycling, and that NDD-associated variants differentially impact these pre- and post-synaptic roles.

Heterozygosity for Trio variants in mice yields phenotypes similar to those observed in NDDs

Individuals with mutations in TRIO present with a range of neurodevelopmental disorder-associated clinical features, including varying degrees of intellectual disability, altered head size, skeletal and facial features, and behavioral abnormalities (23, 25–28, 33, 69). Patients with missense or truncating variants in TRIO that reduce GEF1 activity have mild developmental delay and microcephaly (25, 26, 28, 33). Similarly, we found that heterozygosity for the GEF1-deficient K1431M missense or the K1918X nonsense variants significantly reduced brain weight and/or head size compared to WT mice, along with multiple behavioral impairments. Notably, while both showed impaired motor coordination and learning, only mice bearing the ASD-associated K1431M allele exhibited social interaction deficits.

Both +/K1431M and +/K1918X adult mice had reduced brain-to-body weight ratios compared to WT, but these were driven by different factors. The smaller brain size in +/K1918X mice was associated with a reduction in neuropil and reduced cortical thickness, similar to mice bearing excitatory neuron-specific ablation of one Trio allele (19) and paralleling the reduced gray matter volume and cortical thickness in patients with schizophrenia (70–74). Meanwhile, +/K1431M mice exhibited an overall increase in body weight leading to relatively decreased head width- and brain-to-body weight ratios in these mice. Adult +/K1431M mice had no change in neuropil or cross-sectional brain area, consistent with a prior study describing normal brain size in +/K1431M and K1431M/K1431M mice at E14.5 (20). Rac1 mediates glucose-stimulated insulin secretion from pancreatic islet beta-cells (75–80), which may explain how chronic alterations in Rac1 activity contribute to weight changes in Trio +/K1431M and +/K1918X mice. Most studies of Trio variants have focused on neuronal effects, but expression of Trio in other tissues could explain the increased body weight in these mice, as well as the musculoskeletal abnormalities associated with TRIO variation in humans (23, 25, 27, 32).

Trio variants differentially impact dendritic arbor structure

Rac1 and RhoA signaling is critical for dendrite development (10–13, 81). We found relatively subtle effects of Rac1/RhoA-altering Trio variants on cortical L5 PN dendrites. The reductions in dendritic arborization and length in +/K1918X neurons are consistent with reports of reduced gray matter volume and dendrite alterations in individuals with schizophrenia versus controls (82, 83), but were modest compared to NEX-Trio+/flmice lacking one copy of Trio in excitatory neurons (19). Reduced Trio function in other cell types, such as in inhibitory neurons or glia, or in neurons from other brain regions that project to cortical M1 L5 PNs, may ameliorate the phenotypes in excitatory neurons of +/K1918X relative to NEX-Trio+/fl mice. In addition, changes to L5 PN dendrites in Trio variant mice appeared to be regionally selective within the arbor: +/K1431M neurons had increased arborization only in proximal basal dendrites, while +/M2145T neurons had decreased arborization only in the most distal apical dendrites. These differences may reflect the differential spatiotemporal influence of Trio in regulating Rac1 versus RhoA activity. Alternatively, given our finding for presynaptic roles for Trio, these differences may reflect differential effects of the Trio variants on both excitatory and inhibitory afferent synaptic inputs which play critical roles in shaping the apical and basal dendrites of L5 PNs (84).

Trio variants impact brain Rho GTPase signaling

We show here that the K1431M variant significantly reduces TRIO GEF1 nucleotide exchange on Rac1 in vitro, consistent with previous reports (24, 29). Decreased Rac1 activity was observed in +/K1431M mice at embryonic day 14.5 in the ganglionic eminence, which is enriched for pre-migratory GABAergic interneurons (20). In contrast, we show an increase in active Rac1 levels in +/K1431M postnatal brains (at P0, P21, P42) and synaptosomes (at P42) from cortex, primarily composed of differentiated excitatory neurons. This increase aligns with observed phenotypes (e.g. reduced Pr) and reveals why reductions in brain volume, dendrites and spines, or AMPAR signaling anticipated from reduced Rac1 activity (85–89) were not observed in +/K1431M mice. We propose that the increased Rac1 activity we observed reflects homeostatic compensation in Rac1 regulation occurring between birth to adult ages, and identify changes in Rac1-specific GEFs, e.g. Tiam1 and Vav2, that may contribute to this compensation. Importantly, we did not find changes in Kalirin levels in the adult brain of these Trio mutant mice, suggesting that Kalirin does not compensate for loss of Trio GEF1 activity at this age.

We observed a significant reduction in active RhoA only in purified synaptosomes of +/M2145T brains, reflecting the synaptic compartment as a key locus of Trio function. In addition, despite +/K1918X mice having half the WT levels of Trio protein, we measured little to no change in active Rac1 and RhoA levels in +/K1918X brains. These findings are consistent with recent evidence that the spatiotemporally precise balance of Rac/Rho activity rather than absolute activity levels can mediate cytoskeletal rearrangements (7, 9, 90, 91). Alterations in the activity of additional potential TRIO substrates, such as RhoG (92), Cdc42 (6), and the neurodevelopmentally critical Rac3 (93–95) could also contribute phenotypes in these mice.

The NDD-associated Trio variants cause synaptic transmission, plasticity and excitatory/inhibitory imbalance

Overexpression of a TRIO K1431M variant with reduced GEF1 activity decreased AMPAR-mediated mESPC amplitudes in rat organotypic slices (21), while Rac1 activation increased AMPAR amplitudes by promoting synaptic AMPAR clustering (89). In +/K1431M mice we observed increased AMPAR current amplitudes without significant changes in NMDAR ESPC amplitudes in evoked and miniature EPSCs recordings, consistent with the increase in active Rac1 levels measured in synaptosomes of adult +/K1431M mice. AMPAR EPSC amplitudes were also increased in L5 PNs of +/K1918X mice, which have decreased Rac1 activity at P0 and normalized levels of Rac1 by P21 and P42 compared to WT. This temporal increase of Rac1 activity could be enough to drive a subtle but significant increase in AMPAR tone in +/K1918X.

mIPSC frequencies were decreased in +/K1431M and +/M2145T L5 PNs, while mIPSC amplitude was increased in +/K1918X slices. Sun et al. noted a similar deficit in inhibitory function in +/K1431M prefrontal cortex correlated with reduced interneuron migration to this region, including parvalbumin positive (PV+) neurons (20). In contrast, we did not observe reduced PV+ interneuron numbers in the motor cortex in any Trio variant mice, suggesting that Trio variants may impact the number of inhibitory synapses or transmission properties rather than solely interneuron migration. Overall, heterozygosity for Trio variants dysregulates excitatory and inhibitory synaptic transmission in different patterns, resulting in E/I imbalance, a known driver of NDD phenotypes.

Trio-deficient excitatory neurons are unable to undergo long-term potentiation (LTP) in mouse brain slices (19), which is crucial for working memory in mammals. During LTP ability of Rac1 suggested to be transiently activated and deactivated to regulate AMPAR (96–98). Both +/K1431M and +/K1918X L5 PNs exhibited reduced LTP induction and maintenance. The increased AMPAR resulting from already elevated Rac1 activity (+/K1431M mice) or the inability of reduced levels Trio levels to activate Rac1 (+/K1918X mice) may preclude LTP in these mice. +/M2145T mice showed a striking increase in the induction and maintenance of LTP, correlating with an increased glutamate Pr and SV pool size. The function of RhoA function in plasticity is unknown, but the decrease in RhoA activity measured in cortical synaptosomes may underlie the increase in LTP.

Trio GEF1 and GEF2-deficient variants lead to opposing defects in synaptic release of glutamate

In neuroendocrine, pancreatic beta, and mast cells, Trio GEF1 activity, Rac1 and RhoA are required for regulation of exocytosis (99–103). Recent work suggest that Trio GEF1 can act through Rac1 to regulate presynaptic processes – Rac1 colocalizes with SVs in the axonal boutons to negatively regulate action potential-dependent (synchronous) glutamate Pr and SV replenishment (15, 44, 59, 104). Using the Rac1 inhibitor NSC, we demonstrate that elevated Rac1 activity, possibly due to Tiam1 upregulation, underlies the reduction in synchronous glutamate Pr and SV replenishment without affecting RRP in +/K1431M mice.

Recent work on Rac1 knockout models shows conflicting results regarding their impact on spontaneous (action potential-independent) glutamate Pr, which is critical in synaptogenesis and plasticity (105, 106). Knockout of Rac1 at the Drosophila neuromuscular junction or in cultured mouse hippocampal neurons does not affect the frequency of spontaneous release (15, 59). However, conditional knock out of Rac1 in the Calyx of Held increased spontaneous Pr (44), aligning with the hypothesis that Rac1 activity negatively regulates neurotransmitter release by increasing assembly of actin filaments at the AZ that impede SV fusion. A significant decrease in NMDAR mEPSC frequency in +/K1431M mice suggests defects in spontaneous glutamate Pr in cortical synapses, consistent with increased Rac1 activity, while the increase in miniature AMPAR amplitudes could mask the decrease in AMPAR frequency. This relative impact of Rac1 on spontaneous release may reflect functional differences between these diverse synapse types.

RhoA activity is significantly reduced in cortical synaptosomes of +/M2145T mice, and this is associated with increased RRP size. While a specific function for RhoA in regulating presynaptic release is currently unclear, levels of Munc18, Syp, and Syt3, are all increased in +/M2145T mice and may contribute to the enhanced Pr and altered SV cycling. Taken together, we demonstrate that Trio GEF1 and GEF2 activities play crucial, distinct roles in the presynaptic SV cycle.

Conclusions

TRIO is a risk gene for several NDDs with different patterns of variants observed in different disorders. We show here that variants in Trio that lead to impaired Trio levels or GEF function cause both shared and distinct defects in behavior, neuroanatomy, and synaptic function that reflect variant-specific NDD clinical phenotypes. Our data also demonstrate, for the first time, the differential impact of distinct Trio lesions on glutamate Pr and SVs replenishment, along with alterations in presynaptic release machinery that contribute to these deficits. We propose that distinct Trio variants act via diverse mechanisms to disrupt normal brain function at the synaptic, circuit, and behavior level.

eLife Assessment

This study provides useful findings about the effects of heterozygosity for Trio variants linked to neurodevelopmental and psychiatric disorders in mice. However, the strength of the evidence is limited and incomplete mainly because the experimental flow is difficult to follow, raising concerns about the conclusions' robustness. Clearer connections between variables, such as sex, age, behavior, brain regions, and synaptic measures, and more methodological detail on breeding strategies, test timelines, electrophysiology, and analysis, are needed to support their claims.

Reviewer #1 (Public review):

Summary:

This study explores how heterozygosity for specific neurodevelopmental disorder-associated Trio variants affects mouse behavior, brain structure, and synaptic function, revealing distinct impacts on motor, social, and cognitive behaviors linked to clinical phenotypes. Findings demonstrate that Trio variants yield unique changes in synaptic plasticity and glutamate release, highlighting Trio's critical role in presynaptic function and the importance of examining variant heterozygosity in vivo.

Strengths:

This study generated multiple mouse lines to model each Trio variant, reflecting point mutations observed in human patients with developmental disorders. The authors employed various approaches to evaluate the resulting behavioral, neuronal morphology, synaptic function, and proteomic phenotypes.

Weaknesses:

While the authors present extensive results, the flow of experiments is challenging to follow, raising concerns about the strength of the experimental conclusions. Additionally, the connection between sex, age, behavioral data, brain regions, synaptic transmission, and plasticity lacks clarity, making it difficult to understand the rationale behind each experiment. Clearer explanations of the purpose and connections between experiments are recommended. Furthermore, the methodology requires more detail, particularly regarding mouse breeding strategies, timelines for behavioral tests, electrophysiology conditions, and data analysis procedures.

Reviewer #2 (Public review):

Summary:

The authors generated three mouse lines harboring ASD, Schizophrenia, and Bipolar-associated variants in the TRIO gene. Anatomical, behavioral, physiological, and biochemical assays were deployed to compare and contrast the impact of these mutations in these animals. In this undertaking, the authors sought to identify and characterize the cellular and molecular mechanisms responsible for ASD, Schizophrenia, and Bipolar disorder development.

Strengths:

The establishment of TRIO dysfunction in the development of ASD, Schizophrenia, and Bipolar disorder is very recent and of great interest. Disorder-specific variants have been identified in the TRIO gene, and this study is the first to compare and contrast the impact of these variants in vivo in preclinical models. The impact of these mutations was carefully examined using an impressive host of methods. The authors achieved their goal of identifying behavioral, physiological, and molecular alterations that are disorder/variant specific. The impact of this work is extremely high given the growing appreciation of TRIO dysfunction in a large number of brain-related disorders. This work is very interesting in that it begins to identify the unique and subtle ways brain function is altered in ASD, Schizophrenia, and Bipolar disorder.

Weaknesses:

(1) Most assays were performed in older animals and perhaps only capture alterations that result from homeostatic changes resulting from prodromal pathology that may look very different.

(2) Identification of upregulated (potentially compensating) genes in response to these disorder-specific Trio variants is extremely interesting. However, a functional demonstration of compensation is not provided.

(3) There are instances where data is not shown in the manuscript. See "data not shown". All data collected should be provided even if significant differences are not observed.

I consider weaknesses 1 and 2 minor. While they would very interesting to explore, these experiments might be more appropriate for a follow-up study. I would recommend that the missing data in 3 should be provided in the supplemental material.

Author response:

The following is the authors’ response to the original reviews

eLife Assessment

This study provides useful findings about the effects of heterozygosity for Trio variants linked to neurodevelopmental and psychiatric disorders in mice. However, the strength of the evidence is limited and incomplete mainly because the experimental flow is difficult to follow, raising concerns about the conclusions' robustness. Clearer connections between variables, such as sex, age, behavior, brain regions, and synaptic measures, and more methodological detail on breeding strategies, test timelines, electrophysiology, and analysis, are needed to support their claims.

We appreciate the opportunity to address the constructive feedback provided by eLife and the reviewers. Below, we respond to the overall assessment and individual reviewers' comments, clarifying our experimental approach, addressing concerns, and providing additional details where necessary.

We thank the editors for highlighting the significance of our findings regarding the effects of Trio variant heterozygosity in mice. We acknowledge the feedback concerning the experimental flow and agree that clarity is paramount. To address these concerns:

(1) Connections between variables: The word limit of the initial submission constrained our ability to provide adequate details and connections between variables. We have revised the manuscript to explicitly outline and extend explanations and the relationships between sex, age, behavior, brain regions, and synaptic measures, ensuring that the rationale for each experiment and its relevance to the overall conclusions are improved.

(2) Methodological details: The Methods section of our initial submission was condensed, with key details provided in the Supplemental Methods section. We have merged all into an extended section to improve clarity. We have expanded our description of breeding strategies, test timelines, electrophysiological protocols, and data analysis methods in the revised Methods section. We believe the additions have enhanced the transparency and reproducibility of our study and ensured full support of our conclusions.

(3) Experimental flow: We have revised and extended our results, methods, and discussion sections to clarify the rationale and experimental design to guide readers through the experimental sequence and rationale.

We are confident these revisions address the concerns raised and enhance the robustness and coherence of our findings.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This study explores how heterozygosity for specific neurodevelopmental disorder-associated Trio variants affects mouse behavior, brain structure, and synaptic function, revealing distinct impacts on motor, social, and cognitive behaviors linked to clinical phenotypes. Findings demonstrate that Trio variants yield unique changes in synaptic plasticity and glutamate release, highlighting Trio's critical role in presynaptic function and the importance of examining variant heterozygosity in vivo.

Strengths:

This study generated multiple mouse lines to model each Trio variant, reflecting point mutations observed in human patients with developmental disorders. The authors employed various approaches to evaluate the resulting behavioral, neuronal morphology, synaptic function, and proteomic phenotypes.

Weaknesses:

While the authors present extensive results, the flow of experiments is challenging to follow, raising concerns about the strength of the experimental conclusions. Additionally, the connection between sex, age, behavioral data, brain regions, synaptic transmission, and plasticity lacks clarity, making it difficult to understand the rationale behind each experiment. Clearer explanations of the purpose and connections between experiments are recommended. Furthermore, the methodology requires more detail, particularly regarding mouse breeding strategies, timelines for behavioral tests, electrophysiology conditions, and data analysis procedures.

We appreciate the reviewer’s recognition of the novelty and comprehensiveness of our approach, particularly the generation of multiple mouse lines and our efforts to model Trio variant effects in vivo.

Weaknesses

(1) Experimental flow and rationale and connection between variables: We have expanded on the connections between behavioral data, neuronal morphology, synaptic function, and proteomics in the Results and Discussion sections to clarify how each experiment informs the reasoning and the conclusions and to highlight the relationships between sex, age, behavior, and synaptic measures.

(2) Methodological details: Our initial Methods section was formatted to be short to fulfill word limits on the submitted version, with additional details provided in the Supplemental Methods section. We have merged our Methods and Supplemental Methods sections and expanded on our breeding strategies, test timelines, electrophysiological protocols, and data analysis. We believe these additions enhance the transparency and reproducibility of our study.

(3) Recommendations for the authors: We thank Reviewer #1 for providing several recommendations to improve our manuscript. We have addressed their comments in the revision, as detailed below, adding key experiments that bolster our findings.

Reviewer #2 (Public review):

Summary:

The authors generated three mouse lines harboring ASD, Schizophrenia, and Bipolar-associated variants in the TRIO gene. Anatomical, behavioral, physiological, and biochemical assays were deployed to compare and contrast the impact of these mutations in these animals. In this undertaking, the authors sought to identify and characterize the cellular and molecular mechanisms responsible for ASD, Schizophrenia, and Bipolar disorder development.

Strengths:

The establishment of TRIO dysfunction in the development of ASD, Schizophrenia, and Bipolar disorder is very recent and of great interest. Disorder-specific variants have been identified in the TRIO gene, and this study is the first to compare and contrast the impact of these variants in vivo in preclinical models. The impact of these mutations was carefully examined using an impressive host of methods. The authors achieved their goal of identifying behavioral, physiological, and molecular alterations that are disorder/variant specific. The impact of this work is extremely high given the growing appreciation of TRIO dysfunction in a large number of brain-related disorders. This work is very interesting in that it begins to identify the unique and subtle ways brain function is altered in ASD, Schizophrenia, and Bipolar disorder.

Weaknesses:

(1) Most assays were performed in older animals and perhaps only capture alterations that result from homeostatic changes resulting from prodromal pathology that may look very different.

(2) Identification of upregulated (potentially compensating) genes in response to these disorder-specific Trio variants is extremely interesting. However, a functional demonstration of compensation is not provided.

(3) There are instances where data is not shown in the manuscript. See "data not shown". All data collected should be provided even if significant differences are not observed.

I consider weaknesses 1 and 2 minor. While they would be very interesting to explore, these experiments might be more appropriate for a follow-up study. I would recommend that the missing data in 3 should be provided in the supplemental material.

We are grateful for the reviewer’s recognition of our study’s significance and methodological rigor. The acknowledgment of Trio dysfunction as a novel and impactful area of research is deeply appreciated.

Weaknesses:

We agree that focusing on older animals limits insights into early-stage pathophysiology. However, our goal in this study was to examine the functional impacts of Trio heterozygosity at an adolescent stage and to reveal the ultimate impact of these alleles on synaptic function. Our choice of age aligns with our objectives. Future studies of earlier developmental stages will be beneficial and complement these findings.

Functional compensation:

We tested functional compensation through rescue experiments in +/K1431M brain slices using a Rac1-specific inhibitor, NSC23766, which prevents Rac1 activation by Trio or Tiam1. Our finding that direct Rac1 inhibition normalizes deficient neurotransmitter release in +/K1431M mice strongly suggests that increased Rac1 activity drives this phenotype.

Data not shown:

We will incorporate all previously shown data into the Supplemental Materials, even when results are nonsignificant. We agree that this ensures full transparency and facilitates a more comprehensive evaluation of our findings.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) In Figure 1K-N, the lack of observed differences in +/M2145T mice across all tests raises questions about its validity as a BPD model. Furthermore, the differences in female behavior data compared to males, as shown in the Supplemental section, lack clarification-specifically, whether these variations are due to sex differences or sample size disparities, which is not discussed. Additionally, it's unclear if the same mice were used in tests K through L-N, as the reported numbers differ without explanation; if relevant, any mortality should be reported. Given the observed body weight differences, it is important to display locomotor data, despite the mention of no change in open field results. Lastly, a detailed breeding strategy and timeline for behavioral testing would enhance clarity.

We thank Reviewer 1 for recognizing these confusing points in our behavioral data and seek to add clarification in our Revision as below:

(a) We have revised the text to emphasize our goal to evaluate the impact of NDD-related Trio alleles that have discrete and measurable effects on brain development and function, and not to model specific NDDs (e.g. ASD, SCZ, or BPD). The three specific Trio mutations were chosen based on strong evidence of these mutations impairing the biochemical functions of Trio. We reasoned our approach would reveal how impairing Trio in different ways – i.e. altering protein level or GEF1/GEF2 function – and under genetic conditions (heterozygosity) that mimic those found in individuals with Trio-related disorders impacts brain development and function. The lack of behavioral phenotypes in +/M2145T mice is indeed intriguing, especially given the alterations in electrophysiology and biochemistry experiments. It remains possible that further behavioral analyses of these mice will reveal behavioral phenotypes.

(b) Given that the prevalence and clinical presentation of individuals with various NDDs are influenced by sex, it is possible that the behavioral differences we see in male versus female Trio variant mice reflect human sex difference phenotypes. We have reorganized the Figure panels to clarify these sex differences in behaviors (new Fig. 2, Supp. Fig. 2). We focused on the most significant behavioral phenotypes shared by both sexes in the main text, or in males alone, as our anatomical and electrophysiological experiments were restricted to males to reduce variation due to estrus. The observed behavioral sex differences are not likely due to sample size disparities as power analyses were performed for all experimental results to ensure adequate sample size. A comprehensive study of the mechanisms underlying these behavioral findings merits examination but is outside the scope of this study.

(c) All mice were subjected to all behavioral tests described. No sudden mortality was observed during the behavioral experiments. Outliers in post-hoc statistical analyses were removed, which explains the apparent sample size differences between behavioral tests. We have revised the Data analysis section in our Methods to include these details (Lines 216-289, 450-457).

(d) Results of the open field test have been added to the Supplemental Data (new Supp. Fig. 2) and Results (Lines 532-537)

(e) The Methods section was expanded to include more detail on the breeding strategy (Lines 98-106). A timeline for behavioral testing has also been included in the Figures to enhance clarity (new Fig. 2A).

(2) In Figure 2A-E, head width and brain weight showed significant differences, but not body weight, how come the ratio does not change? Comparing with female results in Supplementary Figure 2A-E, it does show a difference between males and females. It is essential to clarify which sex authors use in all follow-up experiments, including synapse, transmission, and plasticity. Since the males and females have different phenotypes, why do the authors focus on males only? The E plot has no data points on the bar graph. In Figure 2I, it lacks example images for all four conditions.

We greatly appreciate this Reviewer’s attention to details in our brain and body weight data and revised the manuscript to address these concerns.

(a) The ratios of head width/body weight were calculated for each individual mouse. Hence the distribution of the ratio data (old Fig. 2D; new Fig. 3D) differs from the distribution of head width or body weight data alone (old Fig. 2A, 2C, resp.; now Fig. 3A, 3C), and therefore can affect the p-value for statistical significance. The body weight of +/M2145T males is 21.217 ±0.327 g, while for WT males is 21.745 ±0.224 g, a non-significant decrease of 0.528 g (adjusted p=0.3806). These values have been added to the Fig 3. figure legend (Lines 1020-1034) for clarity.

(b) Similar to the behavioral experiments in comment (1), we observed sex differences in head width, brain weight, and body weight in Trio heterozygous variant mice compared to WT counterparts. The differences in the ratios of head width/body weight or brain weight/body weight were the same for both males and females (i.e. head width/body weight ratio is decreased in +/K1431M mice compared to WT regardless of sex, and brain weight/body weight ratio is decreased in both +/K1431M and +/K1918X mice compared to WT regardless of sex). These findings affirm the impact of Trio mutations on these phenotypes across both sexes. We have modified the text to draw more attention to this key point (Lines 554-566 and 777-801).

(c) All experiments (excluding behavior and weight data) were performed in males only to minimize the variation in spine and synapse morphology and physiological activity that can occur due to estrus. We have clarified this in the ‘Animal Work’ section of the Methods (Lines 103-106) as well as in the Figure Legends.

(d) We thank the Reviewer for pointing out Fig. 3E lacks individual data points on the bar graph. Fig. 3E has been modified to now include the brain weight/body weight ratio for each individual mouse rather than across the population, to be consistent with the calculation of head width/body weight ratio (see point 2a).

On original submission, only a representative WT image was selected due to space constraints. The figure (new Fig. 3H and 3K) and figure legend have been revised to include representative traces for all genotypes examined.

(3) In lines 315-320, "None of the Trio variant heterozygotes exhibited altered dendritic spine density on M1 L5 pyramidal neurons compared to WT mice on either apical or basal arbors (Supplementary Figure 3L, M). Electron microscopy of cortical area M1 L5 revealed that synapse density was significantly increased in +/K1918X mice compared to WT (Figure 3A, B), possibly due to a net reduction in neuropil resulting from smaller dendritic arbors." The proposed explanation does not adequately address the observed discrepancy between spine density and synapse density reported in these two experiments. A more thorough analysis is needed to reconcile these conflicting findings and clarify how these distinct measurements may relate to each other in the context of the study's conclusions.

We acknowledge the apparent discrepancy between our dendritic spine density data, which is unchanged from WT for all three Trio variant heterozygotes, and our synapse density data, which showed an increase in +/K1918X M1 L5 compared to WT. We have expanded the explanation for this discrepancy below and added this to the Discussion (Lines 802-811):

a) Because spine density can vary by dendritic branch order and distance from the soma, only protrusions from secondary dendritic arbors of M1 L5 pyramidal neurons were quantified for consistency in analyses. However, all synapses meeting criteria were quantified in EM images, regardless of where they were located along an individual neuron’s arbors. It is possible that the density and distribution of spines along other arbors are different between genotypes but was not captured in our current data.

b) +/K1918X L5 pyramidal neurons are smaller and less complex than WT neurons, especially in the basal compartment corresponding to L5 where EM images were obtained, consistent with the smaller brain size and reduced cortical thickness of +/K1918X mice. We posit that due to their smaller dendritic field size, L5 neurons pack more densely contributing to the increased synapse density observed in +/K1918X M1 L5 cortex. Consistent with this hypothesis, we observed a trend toward increased DAPI+ cell density in M1 L5 of +/K1918X neurons (Supp. Fig. 3N).

(4) In Figure 4, one potential rationale for measuring AMPAR mEPSC frequency is to infer synapse density changes. However, the findings show no frequency change in +/K1431M and +/K1918X, with an increase only in +/M2145T, which contradicts Figure 3 results indicating a trend toward increased density across variants.

This inconsistency is confusing, especially since the authors claim to follow the methodology from the study "Trio Haploinsufficiency Causes Neurodevelopmental Disease-Associated Deficits"; yet, the observed mEPSC amplitude differs significantly from that study, while the frequency remains unaffected. Additionally, the NMDAR mEPSCs reflect combined AMPAR and NMDAR responses at positive holding potentials, with peak amplitude dominated by AMPAR. This inconsistency between holding potential results is unclear, as frequency should theoretically align across negative and positive potentials. For accurate NMDAR mEPSC measurement, it would be optimal to assess amplitude 50 ms post-initial peak and, if possible, increase the holding potential to enhance the driving force given the typically low signal of NMDAR response.

We thank the Reviewer for highlighting these important points.

a) Previous work from our lab and others demonstrate that Trio regulates synaptic AMPA receptor levels, which is why we chose to focus on AMPAR-mediated evoked and miniature EPSC frequencies and amplitudes in the current study. We acknowledge Reviewer 1’s comment on seemingly contradictory results regarding AMPAR mEPSC frequency and synapse density; however, the unchanged AMPAR mEPSC frequency in +/K1431M and +/K1918X mice is consistent with our finding of unaltered dendritic spine density in these mice compared to WT (Supp. Fig. 4L,M). The differences between dendritic spine counts and synapse density is addressed in Response (3) above.

b) While synapse density changes can be inferred from AMPAR mEPSC frequency, mEPSCs are also measures of spontaneous neurotransmitter release changes especially in the absence of changes in synaptic numbers. Notably, the increased mEPSC frequency in the +/M2145T variant is linked to enhanced spontaneous release, not to spine or synapse density changes. These findings are reinforced by increase in counts of synaptic vesicles, calculated PPR changes, and estimates of the Pr and RRP from HFS train analysis. We have included these points in the Discussion (Lines 861-863).

c) While it is tempting to compare the current study to our previously published conditional Trio haploinsufficiency model, we highlight key distinctions that may underlie phenotypic differences between these two mouse models. First, our prior model used a NEX-Cre transgene to ablate one Trio allele from excitatory neurons only beginning at embryonic day 11. In contrast, our Trio variants are expressed in all cell types throughout development, akin to the genetic variants found in individuals with TRIO-related disorders. Second, the Trio variant mice in this study are on a C57BL/6 background, while the Trio haploinsufficient mice were on a mixed 129Sv/J X C57BL/6 background. These differences in the current study may explain why some measures, such as mEPSC amplitude, may not align with those from the Trio conditional haploinsufficiency model.

d) Recordings were performed using specific inhibitors to isolate AMPA and NMDA mEPSCs; these missing methodological details have now been clarified in the updated Methods section (Lines 353-360).

(5) In Supplementary Figure 4, the sample traces indicate a higher NMDA/AMPA ratio, raising the question of whether the AMPA EPSC amplitude changes, as this could reflect PSD length. In Figure 4B, the increased AMPAR mEPSC amplitude in the +/K1918X condition compared to WT suggests an enhanced postsynaptic response, yet the PSD length is reduced in Figure 3C. Can the authors provide a potential hypothesis to explain this?

We appreciate the Reviewer’s feedback. Yes, both evoked and miniature recordings indicate increased AMPAR amplitudes in the +/K1918X variants compared to WT. While PSD length is often linked to synaptic strength, the observed reduction in PSD length in EM PSD length reduction in +/K1918X synapses is small (~6% of WT) and clearly does not correlate with significant changes in synaptic strength. We also note that the whole cell recordings of mEPSCs represent input from all active synapses on the neuron, while PSD length is measured only in synapses of the L5.

(6) In Figure 4, synaptic plasticity appears to decrease to around 50% of baseline; could this reduction be attributed to LTD, or might it result from changes in pipette resistance? Additionally, is the observed potentiation due to changes in presynaptic release probability? Measuring paired-pulse ratio (PPR) before and after induction would clarify this aspect.

We thank the Reviewer for highlighting these important points.

a) We used a well-established theta burst stimulation method for LTP induction in M1 L5 pyramidal neurons. This protocol reliably evokes LTP in WT neurons, as shown in Fig. 5J and K. Both +/K1431M and +/K1918X variants exhibit a slight but discernible increase in evoked excitatory postsynaptic currents (eEPSCs), indicative of the initiation of LTP. Although this increase is smaller compared to WT, the presence of potentiation indicates that long-term depression (LTD) is an unlikely explanation for the observed reduction.

b) To rule out the influence of technical artifacts, pipette resistance was carefully monitored before and after LTP induction. Any cells exhibiting resistance changes exceeding 20% during electrophysiological recordings were excluded from the analysis, ensuring that fluctuations in pipette resistance did not confound LTP measurements. These technical details are denoted in the Methods (Lines 344-346 and 364-366).

c) The potentiation in the +/M2145T variant may stem from increased release probability (Pr) and greater synaptic vesicle availability, but is beyond the scope of this work. We agree this is an intriguing question, not only for +/M2145T but also for +/K1431M mice. Future studies should address this, ideally using models where the Trio variant is selectively introduced into the presynaptic neuron.

(7) In lines 377-380, "The +/M2145T PPR curve was unusual, with significantly reduced PPF at short ISIs, yet clearly increased PPF at longer ISI (Figure 5A, B) compared to WT." The unusual PPR observed at the 100 ms ISI appears unexpected. Can the authors provide an explanation for this anomaly? This finding could suggest atypical presynaptic dynamics or modulation at this specific interval, which may differ from typical synaptic behavior. Further insights into possible mechanisms or experimental conditions affecting this result would be valuable.

"The decreased PPF at initial ISI in +/M2145T mice correlated with increased mEPSC frequency (Fig. 4A-C), suggestive of a possible increase in spontaneous glutamate Pr." If this is the case, it raises the question of why the increased PPR at the initial ISI in +/K1431M does not correspond to the result shown in Figure 4C. This discrepancy suggests that factors beyond initial presynaptic release probability might be influencing the observed synaptic response, or that compensatory mechanisms could be affecting PPR and mEPSC frequency differently in this variant. Further clarification on the interplay between these measurements would help resolve this inconsistency.

We appreciate the Reviewer’s critical reading and genuine interest on this phenotype in +/M2145T mice.

a) The unusual shift of the PPR in +/M2145T at ISI 100ms is fascinating and will require significant additional experimentation that lies beyond the scope of this report to address. We propose it results from altered presynaptic regulators, including increased Syt3 and reduced RhoA activity. Notably, Syt3 influences calcium-dependent SV replenishment, which can cause similar PPR defects (Weingarten DJ et al., 2022); this is now included in the Discussion. (Lines 915-918).

Weingarten DJ, Shrestha A, Juda-Nelson K, Kissiwaa SA, Spruston E, Jackman SL. Fast resupply of synaptic vesicles requires synaptotagmin-3. Nature. 2022 Nov;611(7935):320-325. doi: 10.1038/s41586-022-05337-1. Epub 2022 Oct 19. PMID: 36261524.

b) Thank you for raising the concern in clarity of this statement "The decreased PPF at initial ISI in +/M2145T mice correlated with increased mEPSC frequency (Fig. 4A-C), suggestive of a possible increase in spontaneous glutamate Pr." We have edited the sentence to be more clear (Lines 701-703). First, the K1431M and M2145T variants impact different TRIO catalytic activities disrupting distinct GTPase pathways and differentially affecting presynaptic regulators, which can lead to non-overlapping phenotypes. Also, we expand our discussion that +/K1431M variant data suggest increased AMPAR numbers and fewer silent synapses (Lines 850-855), potentially increasing AMPAR mEPSC frequency and masking the expected decrease in spontaneous release (Lines 905-910). Further experiments are needed, ideally using mixed cultures with TRIO variants in presynaptic neurons with synapses on WT neurons, as minimal stimulation variance analysis in slices would be inconclusive due to its reflection of both Pr and silent synapse changes, similar to mEPSC frequency.

(8) In Figure 5, there is no evidence demonstrating that the NSC inhibitor functions specifically in the +/K1431M condition without affecting other conditions. To verify its specificity, the authors should test the NSC inhibitor's effects across other conditions in parallel, including a control group. Additionally, cumulative RRP measurements should be provided for a more comprehensive assessment of the inhibitor's impact on synaptic function.

We appreciate the Reviewer’s feedback.

a) Previous studies have shown that Rac1 activity can bidirectionally regulate synchronous release probability (Pr). We used the Rac1-specific inhibitor NSC23766 (NSC) to test how Rac1 inhibition impacted the neurotransmitter release deficits observed in +/K1431M mice. We also added control experiments testing the impact of NSC on WT slices. These new experiments are now presented in new Fig. 8 of the revised manuscript, with expanded details in the Results (Lines 737-750) and Discussion (Lines 892-900).

b) To estimate Pr and the RRP, we employed the Decay method as described by (Ruiz et al., 2011), which does not rely on cumulative EPSC plots for RRP estimation. This approach was chosen to account for the initial facilitation in these synapses and fits are done using EPSCs plotted against stimulus number. Additional details have been provided in the Methods section (Lines 367-373).

Ruiz R, Cano R, Casañas JJ, Gaffield MA, Betz WJ, Tabares L. Active zones and the readily releasable pool of synaptic vesicles at the neuromuscular junction of the mouse. J Neurosci. 2011 Feb 9;31(6):2000-8. doi: 10.1523/JNEUROSCI.4663-10.2011. PMID: 21307238; PMCID: PMC6633039.

(9) Given the relevance to NDD, specifying the age window of the mice used is crucial. It is confusing that the synaptic function studies were conducted at P42, while the proteomic analysis was performed at P21. Could the authors clarify the rationale behind using different age points for these analyses? Consistency in age selection, or an explanation for this variation, would help in interpreting the developmental relevance of the findings.

P42 was chosen as the age as it represents young adulthood, by which time clinical features will have already presented in individuals with neurodevelopmental disorders. Our prior studies of NEX-Cre Trio-/- mice found significant measurable differences from WT at this age, after neuronal migration, differentiation, synaptogenesis and pruning have occurred. An earlier developmental timepoint, P21, which coincides with juvenile age in mice, was chosen for proteomics studies to identify earlier changes and potentially targetable and modifiable mechanisms that could influence the phenotypes we observed in older mice. The experiments in P42 versus P21 mice were originally two independent lines of investigation that converged in the current study.

Version 2

Abstract

Abstract

Genetic variants in TRIO are associated with neurodevelopmental disorders (NDDs) including schizophrenia (SCZ), autism spectrum disorder (ASD) and intellectual disability. TRIO uses its two guanine nucleotide exchange factor (GEF) domains to activate GTPases (GEF1: Rac1 and RhoG; GEF2: RhoA) that control neuronal development and connectivity. It remains unclear how discrete TRIO variants differentially impact these neurodevelopmental events. Here, we investigate how heterozygosity for NDD-associated Trio variants – +/K1431M (ASD), +/K1918X (SCZ), and +/M2145T (bipolar disorder, BPD) – impact mouse behavior, brain development, and synapse structure and function. Heterozygosity for different Trio variants impacts motor, social, and cognitive behaviors in distinct ways that model clinical phenotypes in humans. Trio variants differentially impact head and brain size, with corresponding changes in dendritic arbors of motor cortex layer 5 pyramidal neurons (M1 L5 PNs). Although neuronal structure was only modestly altered in the Trio variant heterozygotes, we observe significant changes in synaptic function and plasticity. We also identified distinct changes in glutamate synaptic release in +/K1431M and +/M2145T cortico-cortical synapses. The TRIO K1431M GEF1 domain has impaired ability to promote GTP exchange on Rac1, but +/K1431M mice exhibit increased Rac1 activity, associated with increased levels of the Rac1 GEF Tiam1. Acute Rac1 inhibition with NSC23766 rescued glutamate release deficits in +/K1431M variant cortex. Our work reveals that discrete NDD-associated Trio variants yield overlapping but distinct phenotypes in mice, demonstrates an essential role for Trio in presynaptic glutamate release, and underscores the importance of studying the impact of variant heterozygosity in vivo.

Introduction

Neurodevelopmental disorders (NDDs), including autism spectrum disorder (ASD), intellectual disability (ID), developmental delay (DD), schizophrenia (SCZ) and bipolar disorder (BPD), disrupt brain development and function. NDDs share considerable comorbidities and present with overlapping symptoms, including impairments in cognition, behavior, language, social and motor skills, emotions, and learning ability (1). In addition, NDDs share many risk genes, further suggesting shared mechanistic underpinnings and pathology for these disorders (2). A major challenge remains in understanding how diverse risk-conferring variants contribute to the pathophysiology of various NDDs.

TRIO encodes a large cytoskeletal regulatory protein with two guanine nucleotide exchange factor (GEF) domains for Rho family GTPases - GEF1 activates Rac1 and RhoG, while GEF2 activates RhoA (3–5). TRIO relays signals from cell surface receptors, acting on GTPases to coordinate cytoskeletal rearrangements critical for proper neurodevelopment (6–14). Trio knockout mice exhibit decreased survival, skeletal muscle defects, and severe defects in brain development (15–17). Selective ablation of Trio in either excitatory or inhibitory neurons alters their morphology, impairs synaptic signaling, and yields NDD-related behavioral defects (18, 19). Trio deficiency also leads to aberrations in long-term potentiation (LTP), as Trio-Rac1 signaling promotes AMPA receptor trafficking to increase synaptic strength (18, 20, 21).

Damaging de novo mutations and ultra-rare variants in TRIO are enriched in individuals with NDDs (22–29). Interestingly, nonsense variants spread throughout TRIO are enriched in individuals with SCZ (29, 30), whereas pathogenic missense TRIO variants in or surrounding the GEF1 domain are associated with ASD/ID (23, 27, 28). Variants in the TRIO GEF1 domain that decrease Rac1 activity are associated with milder ID and microcephaly, whereas variants in the adjacent spectrin repeat 8 domain that increase Rac1 activity are associated with more severe ID and macrocephaly (24, 25, 31, 32). Rare missense variants in TRIO have also been observed in BPD, epilepsy, and other disorders, but studies to date are underpowered to establish a causal link with these disorders. Given the wide spectrum of TRIO variants associated with different pathological conditions, a fundamental and unresolved question is the mechanisms by which distinct TRIO variants differentially impact normal mammalian brain development and function. NDD-associated TRIO variants affect in vitro GEF1 and GEF2 activities, cell morphology, inhibitory neuron maturation, axon guidance, and synapse function (18–20, 32–34); however, the impact of heterozygosity for distinct TRIO variants in vivo, as found in individuals with NDDs, has been unexplored.

We report here the comprehensive analysis of mice heterozygous for discrete Trio variants impacting Trio functional domains and associated with different NDDs: +/K1431M in the Trio GEF1 domain associated with ASD, +/K1918X leading to nonsense decay associated with SCZ, and +/M2145T in the Trio GEF2 domain, a de novo mutation found in an individual with BPD that impacts GEF2 activity. We show that these distinct Trio NDD-associated variants differentially impair mouse behavior, brain development, and inhibitory and excitatory synaptic transmission. We show, for the first time, that Trio variants that impact GEF1 and GEF2 activity differentially impact presynaptic neurotransmitter release and synaptic vesicle replenishment. In addition, we found that while the K1431M mutation impairs Trio GEF1 activation of Rac1 in vitro, +/K1431M mice exhibit increased levels of active Rac1. This was associated with increased levels of the Rac1 GEF Tiam1, and acute Rac1 inhibition with NSC23766 rescued glutamate release deficits in +/K1431M variant cortex. Together, our data show how discrete heterozygous TRIO variants that differentially impact Trio biochemical activities yield divergent effects on behavior, neurodevelopment, and synaptic transmission.

Materials and methods

Animal work

All animal work was performed in compliance with federal guidelines and approved by the Yale Institutional Animal Care and Use Committee. All mice were maintained on a C57Bl/6 background and housed on a standard 12-hour light/dark cycle. Mice heterozygous for Trio variants K1431M (Trio+/K1431M), K1918X (Trio+/K1918X), or M2145T (Trio+/M2145T) were crossed with WT (Trio+/+) mice to produce heterozygous Trio variant and WT littermates used in all experiments. As the mice were maintained on the same genetic background, analyses of WT mice were pooled and considered one genotype, irrespective of parental Trio variant genotype. Age-matched mice of both sexes were used for behavioral experiments and for brain and body weight measurements. Only male mice were used for electrophysiological and neuroanatomical analyses to reduce potential variation in dendritic spines and activity due to estrus cycle (35).

Generation of Trio mutant mice

Mice heterozygous for Trio variants K1431M, K1918X, or M2145T were generated via CRISPR/Cas-mediated genome editing (36, 37). Potential Cas9 target guide (protospacer) sequences in the vicinity of the Trio K1431, K1918, and M2145 codons were screened using the online tool CRISPOR (38) and candidates were selected. Templates for sgRNA synthesis were generated by PCR from a pX330 template (Addgene), sgRNAs were transcribed in vitro and purified (Megashortscript, MegaClear; ThermoFisher). sgRNA/Cas9 RNPs were complexed and tested for activity by zygote electroporation, incubation of embryos to blastocyst stage, and genotype scoring of indel creation at the target sites. sgRNAs that demonstrated the highest activity were selected for creating the knock-in alleles. Guide primers for generating the template for transcription included a 5’ T7 promoter and a 3’ sgRNA scaffold sequence and were as follows (protospacer sequence capitalized):

K1431M: 5’-tgtaatacgactcactataggACCTGATCAAACCAGTTCAGgttttagagctagaaatagc-3’

K1918X: 5’-tgtaatacgactcactataggAAGCTTCTCACGCACGCAGGgttttagagctagaaatagc-3’

M2145T: 5’-tgtaatacgactcactataggATGACATGATGAACGTCGGGgttttagagctagaaatagc-3’

Recombination template oligos (Integrated DNA Technologies, San Diego CA) were designed to create the desired codon changes, with incorporation of silent mutations to destroy the PAM and prevent sgRNA recognition of the newly created alleles as well as create a new restriction site for genotyping. sgRNA/Cas9 RNP and the template oligo were electroporated into C57Bl/6J (JAX) zygotes (37). Embryos were transferred to the oviducts of pseudopregnant CD-1 foster females using standard techniques (39). Genotype screening of tissue biopsies from founder pups was performed by PCR amplification and Sanger sequencing to identify the desired base changes, followed by backcrossing to C57Bl/6 mice and sequence confirmation to establish germline transmission of the correctly targeted alleles. Recombination template oligos were as follows (silent mutations bolded and italicized; new restriction digest site underlined):

K1431M: 5’-gccaattccatctcttcctACCTGATCAAACCTGTACAGCGTataacaaTgtatcagctccttttaaaggtgtgtat gaaacattgtccatctgcaacatcacaccctgtgttgatatgtatcccct-3’ (creates BsrGI site)

K1918X: 5’-agctgccatctatggcaaaggcatttgtacacaggtgtgagagtgactcgcgaaACTGCTGCATGCGTGAGAA GCTTaccattttgctttAcacgagttcttcaatagcactgacgagctcagctgcgctgggag-3’ (creates SphI site)

M2145T: 5’-agggcccctcaataaacaatgcattacgtcaaatccctgcaGTCGACCAacgttcatcGtgtcattacagcgcttcggca cgatgcacatgacttctacagctttctgtgggtggagagaaagca-3’ (creates SalI site)

Mouse genotyping

Genotypes were determined by PCR of mouse DNA using the following primer pairs:

GFP_For: 5’-gcacgacttcttcaagtccgccatgcc-3’

GFP_Rev: 5’-gcggatcttgaagttcaccttgatgcc-3’

Trio1431_For: 5’-ttgtcattaatgtggtactgtgccc-3’

Trio1431_Rev: 5’-gacaggccaagaaatgtcagtg-3’

Trio1918_For: 5’-tacgagggagttcactgtctg-3’

Trio1918_Rev: 5’-agtgcaggctatgcttcgttta-3’

Trio2145_For: 5’-gcctggacacatccgaattaga-3’

Trio2145_Rev: 5’-aataaccccggacagaggaaag-3’

To distinguish Trio variant and WT alleles, PCR was followed by restriction digest at 37°C for at least 4 hours (BsrGI for K1431M, SphI for K1918X, SalI for M2145T) and subjected to agarose gel electrophoresis.

WT samples resulted in a band at 530 kb for K1431 litters, 550 kb for K1918 litters, 400 kb for M2145 litters; while Trio heterozygous variants resulted in bands at 450 and 530 kb for +/K1431M, 350 and 550 kb for +/K1918X, 350 and 400 kb for +/M2145T. All mice used in experiments were genotyped twice from DNA samples collected at two different time points, prior to weaning and again post-experiment for validation.

In vitro GEF Assays

The K1431M point mutant was generated via site-directed mutagenesis of human WT TRIO GEF1 using the following oligos: 5’-cagcgaataacgatgtatcagctcc-3’ and 5’-ggagctgatacatcgttattcgctg-3’. Recombinant WT and K1431M human TRIO GEF1 and Rac1 proteins were purified from bacteria as previously described (40). GEF activity was monitored by the decrease in fluorescent signal (λexcitation = 488 nm; λemission = 535 nm) as GTP was exchanged for BODIPY-FL-GDP on Rac1 over 30 min, as previously described (32, 40).

Body weight, brain weight, and head width measurements

Ear-to-ear head width of P42 mice was measured with calipers in anesthetized mice. Brain weight was measured after removing the olfactory bulbs and brain stem.

Brain lysate preparation

Whole brain lysates from P0-P1 or P35-P42 mice were prepared as previously described (18, 41, 42), with minor modifications. Cerebella, hippocampi, and cortices were rapidly removed, snap-frozen in liquid nitrogen and stored at -80°C until lysate preparation. Tissue was homogenized in ice-cold RIPA buffer (1% NP-40, 0.1% SDS, 50 mM Tris pH 8, 150 mM NaCl, 0.5% sodium deoxycholate) supplemented with protease and phosphatase inhibitors (Roche 11873580001; Roche 04906837001), then clarified by brief centrifugation; aliquots were snap frozen and stored at -80° C until immunoblotting. For Trio immunoblots, tissue was sonicated in homogenization buffer (1% SDS, 50 mM Tris pH 7.4, 2 mM EDTA), supplemented with protease and phosphatase inhibitors, heated at 95° C for 5 min, then clarified, snap frozen, and stored at -80° C until immunoblotting. Protein concentrations were determined with a bicinchoninic acid (BCA) assay (Pierce).

Cortical synaptosome prep

Crude synaptosomes were prepared as previously described (41, 42). Briefly, cortices were rapidly dissected from anesthetized P39-P42 mice and immediately homogenized in cold buffer (118 mM NaCl, 4.7 mM KCl, 1.2 mM MgSO4, 2.5 mM CaCl2, 1.53 mM KH2PO4, 212.7 mM glucose) supplemented with protease and phosphatase inhibitors (Roche). Homogenates were passed through a series of nylon filters of descending pore size: 40 μm, 10 μm, and finally 5 μm. Samples were centrifuged for 15 min at 1,000 x g at 4°C; the supernatant was discarded, and the pellet was resuspended in an appropriate buffer for G-LISA measurements or immunoblotting. Mice from at least 3 litters per genotype were processed concurrently with WT littermates.

Western blot and quantification

Brain lysates were separated on SDS-PAGE gels, transferred to nitrocellulose membranes, stained with Ponceau S, blocked in 5% nonfat milk in TBS-T, and incubated with primary antibodies overnight at 4°C, then with conjugated secondary antibodies at RT for 1 h (see Key Resources Table for list of antibodies). Images were captured by a ChemiDoc Imaging System (Bio-Rad) and quantified in ImageJ. Signal intensity was normalized to Ponceau S, then to the WT average.

G-LISA

Active GTP-bound GTPase levels in brain lysates were measured using G-LISA activation assay kits for Rac1 (Cytoskeleton, Inc., BK128) and RhoA (Cytoskeleton, Inc., BK124). Brains or crude synaptosome pellets were homogenized in G-LISA lysis buffer supplemented with protease and phosphatase inhibitors (Roche), normalized to the same concentration as determined by Precision Red Advanced Protein Assay, and applied to the G-LISA matrices and processed according to manufacturer’s protocols. Absorbances at 490 nm (OD490) were background-subtracted and normalized to the WT average.

Behavioral tests

Behavioral tests were performed in both male and female mice 6-8 weeks of age (P42-P56). Mice were habituated to handling for 5 minutes per day for 5 days prior to experiments, and habituated in the test facility separate from housing for at least 30 minutes prior to starting the task. We performed the Kondziela inverted screen test for motor strength, accelerating rotarod for motor coordination and learning, social preference test, novel object recognition task, open field test, elevated plus maze, and nestlet shredding, as previously described (18, 43–46). All behavioral studies were performed and analyzed by an experimenter blinded to genotype. Individual data points could be excluded from analysis if a mouse failed to properly acclimate to the testing conditions, as described in detail for specific tests below.

Accelerating rotarod

A five-lane rotarod treadmill (Med Associates ENV-577M 8.75’’ rod circumference) was used for accelerating analyses. Mice were acclimated to the rotating rod at 4 RPM for 10 sec before beginning the test; mice that were unable to stay on the rod during this period were excluded from analyses. The rod then accelerated from 4 to 40 RPM over the course of 5 minutes (1.2 RPM/10 s) before leveling off at 40 RPM for 2.5 minutes. Five trials were performed per mouse with a rest time of at least 10 minutes between trials. The latency to fall was recorded per trial.

Kondziela inverted screen test

The screen apparatus consisted of a 40 cm square of wire mesh composed of 12 mm squares with 1 mm diameter wire, surrounded by a square wooden frame 3.5 cm thick. Mice were placed in the center of the screen and inverted over 2 seconds over a clear open box 45 cm high. During the training session on day 1, mice were habituated to hanging on the screen until a cumulative hang time of 2 minutes was reached; mice were excluded if they were unable to meet this requirement. During the test session on day 2, mice were placed on the screen as before and given 3 trials to reach a maximum inverted hang time of 8 minutes, with 10-minute rest periods between trials. Mice that were unsuccessful at reaching the 8-minute mark were excluded from the accelerating rotarod test.

Open field test

Mice were placed in a large 16x16 inch square clear plexiglass SuperFlex cage and monitored using the Fusion software connected to a light beam array (Omnitech Electronics, Open Field Test, Version 4.5) for 10 minutes. The following measurements were analyzed based upon beam breaks: total distance traveled, ambulatory time (successive beam breaks), and vertical activity time (z axis beam breaks).

Social preference test

Mice were placed in a 50 cm x 50 cm plexiglass box with two wire-mesh pencil holders (open side facing downward) for 10 minutes for habituation. Then, the mouse was briefly removed while an inanimate object (Duplo blocks of a similar size and color to a mouse) and a male conspecific ‘‘stranger’’ mouse were placed underneath separate wire-mesh pencil holders. The test mouse was then returned to the cage for 10 minutes, tracked using the AnyMaze software (Stoelting Co.), and scored for entries into and time within a social target zone and an equivalent nonsocial target zone. The target zones were defined by an annulus around the wire enclosure.

Novel object recognition task

During object familiarization on day 1, two identical objects were placed on the left and right sides of the testing cage. Mice explored the objects until they accumulated 30 s of tactile exploration time, defined as direct oral or nasal contact. Mice were then returned to their home cage for 48 hours. During the novel object recognition task on day 3, one familiar and one novel object were placed on either side of the testing cage, and mice explored the objects until they accumulated 30 s of tactile exploration time. Mice were excluded from analysis if they failed to explore both objects or accumulate 30 s of tactile exploration time within 6 minutes on either day.

Elevated plus maze

The apparatus is raised 36.5 cm above the ground with a 5 cm square center platform and four 30 cm x 5 cm arms; two of the arms are enclosed by 16 cm high walls (Stoelting Co, Wood Dale, IL). Mice were tested in 3 trials that were 5 minutes long with at least 10 minutes to rest between trials. Mice were placed directly on the center platform and allowed to freely explore all arms of the maze, tracked using the AnyMaze software (Stoelting Co.), and scored for time spent in each arm.

Nestlet shredding

A standard polycarbonate mouse cage (19×29×13 cm) was filled with unscented bedding material to a depth of 0.5 cm, the surface was leveled, and a fitted filter-top cover was placed on top. Commercially available cotton fiber nestlets (5 cm x 5 cm, 5 mm thick, 2-2.5 g each) were weighed on an analytical balance before being placed on top of the bedding. The mouse was placed into the test cage without food or water and left undisturbed for 30 minutes. The remaining intact nestlet was removed and allowed to dry overnight. The weight of the dried unshredded nestlet was divided by the starting weight to calculate percentage of nestlet shredded.

Animal perfusion and tissue processing

Mice were transcardially perfused with heparinized PBS followed by 4% paraformaldehyde (PFA) in phosphate buffered saline (PBS). Brains were postfixed in 4% PFA at 4°C for 24 h, then sliced coronally at 30 μm for Nissl stain and immunohistochemistry and at 200 μm for dendritic arbor and spine analysis. Sections were processed as per application and coverslip-mounted with ProLong Diamond Antifade Mountant (P36961, ThermoFisher). For electron microscopy, mice were perfused with 4% PFA supplemented with 2% glutaraldehyde.

Nissl staining

Nissl staining on 30 μm brain slices was performed according to standard protocols (47). Brightfield images were acquired at 20X on a SlideView VS200 slide scanner (Olympus). Coronal sections containing the decussation of the anterior commissure were selected for analysis and quantified using QuPath software. Measurements from three consecutive slices were averaged per mouse.

Immunostaining

Desired sections were identified by comparison to a mouse brain atlas (48). Three 30-μm coronal sections from the motor cortex, spaced at 120-μm intervals and containing the rostral decussation of the corpus callosum were used for immunostaining. Mounted sections were boiled in Na-Citrate buffer (10 mM, pH 6.4) for 45 min, blocked for 1 h in 5% normal goat serum (NGS) in 0.3% PBS-T, incubated overnight at 4°C with primary antibodies (Key Resources Table), then with appropriate AlexaFluor-conjugated secondary antibodies for 1 h at RT, and counterstained for 10 min with DAPI. Controls lacking primary antibody were used to confirm staining specificity. Fluorescent images were acquired at 10X on a SlideView VS200 slide scanner (Olympus). Semi-automatic cell counts for DAPI+, NeuN+, and PV+ stained cells were obtained using QuPath by an experimenter blinded to genotype.

Dendritic arbor reconstructions and dendritic spine density analysis

Mice of all genotypes examined were intercrossed with mice bearing a thy1-GFP transgene (M line) (49). Fluorescent images of neurons contained within 200 μm coronal sections were acquired at 20X for arbor reconstructions (Zeiss LSM900) and at 100x on an UltraVIEW VoX spinning disc confocal microscope for dendritic spines. Dendrites were traced and measured using the Simple Neurite Tracer Fiji plugin (SNT) in ImageJ (NIH). Spines were manually counted from max projection z-stack images on secondary apical and basal dendritic branches of L5 PNs.

Electron microscopy

Electron microscopy of synapses in M1 layer 5 was performed as previously described (18). Asymmetric excitatory synapses were identified as an electron-dense post-synaptic density (PSD) apposed to a presynaptic terminal containing synaptic vesicles. Measurements of synaptic features, including synapse density, PSD length, spine head area, presynaptic bouton area, and synaptic vesicle distribution, were quantified in ImageJ as previously described (50–53), from 5-18 fields of view (∼50 μm2) per mouse by a reviewer blinded to genotype.

Electrophysiological recordings

Acute slices were prepared from mice at P35-42 as previously described (18) with modifications noted here. Coronal slices of M1-M2 cortex were cut at 360 µm in ice-cold N-Methyl-D-glucamine-aCSF (NMDG-aCSF, in mM): 120 NMDG, 2.5 KCl, 7 MgSO4, 1.25 NaH2PO4, 0.5 CaCl2, 28 NaHCO3, 13 glucose, 7 sucrose, saturated with 95% O2-5% O2 at 300-320 mOsmol/L, pH 7.4) on a vibratome (Leica). Slices were recovered in ACSF (in mM): 120 NaCl, 2 KCl, 1.2 NaH2PO4, 26 NaHCO3, 1.3 MgSO4, 1.6 CaCl2, 11 glucose, 2 ascorbic acid, 4 Na-lactic acid, 2 Na-pyruvate saturated with 95% O2-5% O2 at 300-320 mOsmol/L, pH 7.4, at 32°C for 20 min followed by 1 h recovery and recordings at RT. Recorded signals were acquired at 100 kHz sampling rate and low pass filtered at 6 kHz. Cells were excluded from analysis if series resistance changed >20%.

Excitatory postsynaptic currents (EPSCs) were evoked (e) by a tungsten bipolar stimulating electrode placed in M1 L2/3 with stimulus intensities that yielded 40%-50% of the maximal response. AMPAR-eEPSC amplitudes were measured at the peak of the EPSC recorded at -70 mV holding potential. NMDAR-eEPSC amplitudes were recorded from the same neurons at +40 mV and measured at 40 ms after the stimulus artifact when AMPAR-eEPSCs have decayed.

Miniature EPSCs (mEPSCs) were recorded in whole-cell patch mode with 1 μM tetrodotoxin, 5 μM strychnine and 20 μM bicuculline methiodide (BMI) in the external solution. AMPAR-mEPSCs were recorded at -70 mV with 50 μM D-2-Amino-5-phosphonovaleric acid (D-AP5) to block NMDAR. NMDAR-mEPSCs were recorded at +45 mV with 5 μM Cyanquixaline (CNQX) to block AMPAR. Gamma-aminobutyric acid receptor and glycine receptor inhibitory postsynaptic currents (GABAR/GlyR-mIPSCs) were recorded at +15 mV with D-AP5 and CNQX. Amplitude and frequency of miniature events were detected using Mini-Analysis software (Synaptosoft, Decatur, GA) and inspected visually.

Paired-pulse stimulation was applied with interstimulus intervals (ISIs) of 35, 60, 100, 200 or 300 ms. The amplitude of the second eEPSC (EPSC2), evoked by the second of the paired-pulse stimulations, was divided by eEPSC1 amplitude to obtain a paired-pulse ratio (PPR). Long-term potentiation (LTP) of evoked EPSCs (recorded at -70 mV) was induced by theta-burst stimulation (TBS, five trains of four pulses at 100 Hz in 200 ms intervals, repeated four times with an interval of 10 s) in the presence of 5 μM strychnine and 20 μM BMI.

Estimations of the readily releasable pool (RRP) size and synaptic vesicles (SVs) depletion rate were made from recordings of AMPAR-eEPSCs using high-frequency stimulation (HFS, 40Hz) trains of 15 pulses to deplete the RRP. Trains were analyzed to estimate glutamate probability of release (Pr) and RRP for each neuron using the decay method as previously described (54, 55) using OriginLab 10 software. RRP recovery was approximated as fractional recovery using initial train stimulation with interstimulus intervals of 0.1, 2, 5, and 10 s, followed by a single stimulus. Fractional recovery for each ISI was calculated as (eEPSCs2-eEPSCt1ss)/(eEPSCt1-eEPSCt1ss), where eEPSCs2-amplitude of the stimulus applied after the train, eEPSCt1-first amplitude of the train, eEPSCt1ss-steady state amplitude of the train (last 5 responses). 1 mM kynurenic acid was added to prevent AMPA receptor saturation; 50 μM vD-AP5 added to block NMDA receptors, 5 μM strychnine and 20 μM BMI to block GABA and glycine receptors.

PPR and train recordings for the rescue experiments were performed as described above with addition of 100μM NSC23766 into the recording solution with 5 min incubation period to allow for efficient slice penetration.

Quantitative Proteomics - Sample preparation

Cortices from male P21 mice (4 per genotype) was removed and immediately frozen in liquid nitrogen. Mouse brain tissues were transferred to Covaris tissueTUBE TT1 (Part No. 520001, Plug Part No. 520006), kept on dry ice, and flash frozen in liquid nitrogen prior to cryopulverization using a Covaris CP02 cryoPREP Automated Dry Pulverizer (Part No. 500001). The resulting powdered samples were transferred to 1.5 ml microcentrifuge tubes and lysed in 300 μl lysis buffer (8M urea in 50 mM Tris pH 8.0, 150 mM NaCl, 1 mM EDTA and protease phosphatase inhibitor) on ice. Lysates were sonicated using AFA Covaris E220 (Part No. 500239). The sonicated samples were centrifuged for 30 min at 10,000 rpm. Clear supernatants were transferred to new tubes and quantified with BCA assay (Thermo). Lysates were reduced with 10 mM TCEP and alkylated with 10 mM iodoacetamide for 30 min in the dark at room temperature. 1 mg lysates were diluted to 4 M urea with 200 mM Tris pH 8.0 and digested with Lys-C (1:50 protease:protein ratio) for 90 min at 37°C. Samples were then diluted to less than 2 M urea with 200 mM Tris pH 8.0 and digested with trypsin at the same ratio overnight at 37° C. The samples were acidified with 1% formic acid and desalted using 50 mg per 1 cc Sep-Pak (Waters). A 250 μg aliquot of each sample was frozen and dried, then stored at -80° C until tandem mass tag (TMTpro 16, Thermo lot# WI325918) peptide labeling according to manufacturer instructions.

Desalted 16-plex TMT labeled peptides (5 mg total) were subjected to basic reverse-phase (RP) chromatography. Peptides were separated using a 4.6 mm x 250 mm RP Zorbax 300 A Extend-C18 column (Agilent, 5 mm bead size) on an Agilent 1260 series HPLC. The 96 min gradient of solvent A (2% ACN, 5 mM ammonium formate) to solvent B (90% ACN, 5 mM ammonium formate) stayed at 0% B for 7 min, from 0% to 16% B over 6 min, from 16% to 40% over 60 min, from 40% to 44% over 4 min, from 44% to 60% over 5 min and maintained at 60% B over 14 min. Flow rate was 1 ml/min. 90 fractions were collected and concatenated to down to 24 fractions. 5% of each final fraction was used for proteome analysis. The remaining 95% of each of the 25 fractions were concatenated to 13 fractions. These 13 fractions were IMAC enriched using Ni-NTA Agarose beads as described for phosphoproteome analysis (56).

Quantitative Proteomics - Data acquisition

Proteome analysis was performed using an EASY-nLC 1200 UHPLC coupled to an Orbitrap Eclipse mass spectrometer (Thermo Fisher Scientific). The online peptide separation was performed using a 25 cm x 75 mm i.d. silica picofrit capillary column (New Objectives) packed with 1.9 mm ReproSil-Pur C18-AQ beads (Dr. Maisch GmbH). The 110 min method, 84 min effective gradient of solvent A (2% ACN, 0.1% FA) and solvent B (80% ACN, 0.1% FA). The gradient started at 2% B and increased to 10% B over 1 min, from 10% to 50% B over 84 min, from 50% to 72% B over 9 min, from 72% to 90% B over 1 min, stayed at 90% B over 5 min, then dropped to 60% B over 1 min and maintained at 60% B for 9 min. The flow rate was at 200 nl/min. The Eclipse mass spectrometer performed data-dependent acquisition in positive ion mode. MS1 spectra scanned a range of 350 – 2000 m/z at 60000 resolution, with maximum injection time of 50 ms and 100% AGC target. MS2 spectra scanned at first mass of 110 m/z at 50000 resolution. The cycle time for MS2 scan was set for 2 s for charge state of 1 < x < 6, with isolation window of 0.7 m/z. The AGC target for MS2 was 60% with a maximum injection time of 105 ms. An empirically determined normalized collision energy was set to 32. Dynamic exclusion was set to 20 s.

Quantitative Proteomics - Data analysis

All .raw files were searched using Spectrum Mill (Agilent). MS2 spectra were searched against Uniprot Mouse Database (20171228, 47069 mouse entries, 264 common laboratory contaminants) with a mass tolerance of 20 ppm for both the precursor and product ions. The enzyme specificity was set for Trypsin and allowed up to three missed cleavages. The fixed modification was carbamidomethylation at cysteine. TMT labeling was required at lysine, but peptide N-termini were allowed to be either labeled or unlabeled. Allowed variable modifications for whole proteome datasets were acetylation of protein N-termini, oxidized methionine, deamidation of asparagine, and pyroglutamic acid at peptide N-terminal glutamine, with a precursor MH+ shift range of -18 to 64 Da. The false discovery rate was less than 1%. Protein identifications were discarded if the protein was only observed by a single peptide. Protein subgroups were collapsed to the proteoform with the most evidence.

Gene Set Enrichment Analysis

Mouse UniProtIDs for identified proteins were converted to human orthologs and ranked by signed log10(nominal p-value) (with sign indicating direction of fold-change from WT). GSEA 4.3.3 (57, 58) was used to run the GSEA Preranked tests against all gene sets in the Human MSigDB Collection C2 v2023.2. SynGO 1.2 (59) was used to identify synaptic ontologies for differentially expressed proteins with nominal p-value <0.05.

Data Analysis

Statistical analyses were performed using GraphPad Prism 10. Data in bar graphs are presented as mean ± SEM, with individual data points graphed when applicable. Sample size ‘n’ is annotated within the bars or in the figure legend for each group. Distributions were tested for normality and outliers (>1.5x the interquartile distance for behavioral data, or by Prism ROUT method at Q = 5% for other metrics) were removed before proceeding with statistical tests. Specific details of statistical tests performed, adjusted with post-hoc Bonferroni test for multiple comparisons (MC) where appropriate, are included in the figure legends. Significance was defined by a p-value less than 0.05: nsp<0.1; *p<0.05; **p<0.01; ***p<0.001; ****p<0.0001.

Results

Generation of Trio variant mice and impact on Trio isoforms

To evaluate how different ways of impairing Trio could affect mammalian brain development and function, we used CRISPR/Cas9 technology to generate mice bearing heterozygous Trio variant alleles in different Trio functional domains (Fig. 1A,B). We chose these alleles for their discrete and measurable effects on TRIO levels or GEF activity in vitro: K1431M impairs TRIO GEF1 interaction with and subsequent activation of Rac1 in vitro up to 8-fold (Supp. Fig. 1A,B) (23, 28), M2145T TRIO GEF2 has a reduced ability to activate RhoA as a function of protein concentration in cells (23), and K1918X is predicted to lead to nonsense-mediated decay and loss of Trio protein (23). These Trio variant alleles are found heterozygous in individuals with NDDs: +/K1431M is associated with ASD, +/K1918X is associated with SCZ, and +/M2145T is a de novo mutation found in an individual with BPD (60).

Genetically engineered mice with heterozygosity for K1431M, K1918X, or M2145T Trio variants have divergent effects on Trio protein expression and Rho GTPase activity.

(A) Schematic of major Trio isoforms present in the adult mouse brain, with locations of engineered neurodevelopmental disease (NDD)-associated Trio variants: K1431M is a rare missense variant in the GEF1 DH domain associated with autism spectrum disorder (ASD); a K1918X nonsense variant that lies just before the GEF2 domain associated with schizophrenia (SCZ); and M2145T missense variant in the GEF2 DH domain found in an individual with bipolar disorder (BPD). (B) Representative sequencing chromatograms of WT, and Trio variant mice. Arrows indicate heterozygosity for the variant alleles. (C) Representative immunoblots for Trio in P0 brain lysates using an antibody against Trio spectrin repeats (SR5-6). (D) Quantification of Trio protein levels from P0 brain lysates. Trio protein levels are reduced only in the brains of +/K1918X mice compared to WT controls (0.545 ± 0.126 of WT level, p=0.0046). (E-H) Activity levels of Rac1 (E,G) and RhoA (F,H) in whole brain homogenates of neonate (P0, E-F) and adult (P42, G-H) Trio variant mice as measured by G-LISA assay. Rac1 activity is increased in +/K1431M mice relative to WT at both ages (1.106 ± 0.027-fold at P0, p=0.0035; 1.509 ± 0.175-fold at P42, p=0.0279) and decreased in neonate +/K1918X mice (0.908 ± 0.0.032-fold, p=0.0230), with a trend towards increased activity in adult +/M2145T mice (1.438 ± 0.183-fold, p=0.0843); meanwhile RhoA activity appears unchanged in all mice relative to WT, though there may be a trend towards decreased activity in +/K1918X neonates (0.840 ± 0.074-fold, p=0.1292). (I,J) Activity levels of Rac1 (I) and RhoA (J) in synaptosomes isolated from P42 mouse cortex. Rac1 activity is increased in +/K1431M synaptosomes (1.125 ± 0.107-fold, p=0.0023), while RhoA activity is decreased in +/M2145T synaptosomes (0.731 ± 0.042-fold, p=0.0093) relative to WT. All data shown as mean ± SEM. For (D-J), one-way ANOVA with post-hoc Bonferroni MC test identified differences from WT (nsp<0.1, *p<0.05, **p<0.01). Mouse numbers per group are shown in bars.

Because Trio knockout mice exhibit decreased survival (15–17), we first verified the viability of mice bearing these CRISPR/Cas9-generated Trio alleles. Mice heterozygous for any one of these three Trio alleles survived to adulthood. Mice homozygous for Trio K1431M and M2145T survived to adulthood, with genotypes from offspring of +/K1431M intercrosses observed in Mendelian ratios, but fewer than expected M2145T homozygotes obtained from +/M2145T intercrosses (Supp. Fig. 1C). K1918X homozygote pups were not observed (Supp. Fig. 1C), as expected for a Trio null allele (15, 17). We focused on heterozygotes, as most rare damaging Trio variants are heterozygous in humans.

Alternative splicing of the full-length TRIO transcript generates multiple brain-specific isoforms of various sizes, whose expression level varies by brain region and developmental age: cortex-predominant Trio9S (263 kDa) and Trio9L (277 kDa), brain-wide Duet (145 kDa), and cerebellum-specific Trio8 (217 kDa) (Fig. 1A) (61, 62). Heterozygosity for the K1431M and M2145T alleles did not alter the levels of the predominant Trio isoforms in the brain at postnatal day 0 (P0, neonate) (Fig. 1C-D) or at P42 (young adult) (Supp. Fig. 1D-L). Trio9 protein levels were reduced by ∼50% in the brains of +/K1918X mice at P0 (Fig. 1C-D) and P42 (Supp. Fig. 1D-L), and we did not detect the presence of residual truncated protein (expected at 217 kDa), suggesting this K1918X mutation indeed leads to nonsense-mediated decay as predicted. Meanwhile, as the K1918 site is not contained in these isoforms, the levels of Trio8 and Duet were unaffected (Supp. Fig. 1D-L).

Trio variant alleles differentially impact active Rho family GTPase levels

Having demonstrated that heterozygosity for the K1431M and M2145T alleles does not alter Trio protein levels in the brain, we can attribute any observed differences to the effects of these missense mutations on Trio GEF domain function rather than changes in Trio protein expression. Given the effects of these variants on TRIO GEF1/2 activities in vitro, we measured levels of active GTP-bound Rac1 and RhoA in brains of neonatal (P0) and young adult (P42) Trio variant-bearing mice (Fig. 1E-H).

In concordance with the ∼50% reduction in Trio levels in +/K1918X brains (Fig. 1C,D), Rac1 activity was decreased in P0 +/K1918X brains (91% of WT activity, p<0.05) (Fig. 1E), with a trend toward decreased active RhoA levels (84% of WT, p=0.0865) (Fig. 1F). However, by P42, +/K1918X brains did not differ from WT in Rac1 or RhoA activity (Fig. 1G,H), despite the persistent ∼50% reduction in Trio9 protein level at this age (Supp. Fig. 1D-F,H-J).

As K1431M decreases TRIO GEF1 catalytic activity in vitro (Supp. Fig. 1A,B) (19, 23, 28), we anticipated Rac1 activity would also be reduced in +/K1431M mice. Instead, active Rac1 levels were significantly elevated in +/K1431M whole brain lysates compared to WT controls, showing a 111% increase at P0 (p<0.01) (Fig. 1E) and an even greater increase of 150% at P42 (p<0.05) (Fig. 1G). Despite evidence of M2145T impacting TRIO GEF2 catalytic activity in vitro (23), we did not detect changes in active RhoA levels at the whole brain lysate level in +/M2145T mice at either age (Fig. 1F,H).

Because Trio, Rac1, and RhoA are enriched at synapses (14, 21, 23, 63), we also measured GTPase activities in synaptosomes (Supp. Fig. 1M-P), within the same age range as the electrophysiological recordings performed on these mice. Active Rac1 was significantly increased in +/K1431M crude synaptosomes from P42 cortex (112% of WT, p<0.01) (Fig. 1I), consistent with our findings in +/K1431M whole brain lysates (Fig. 1E,G). We also measured decreased levels of active RhoA in P42 +/M2145T cortical synaptosomes compared to WT (73% of WT, p<0.01) (Fig. 1J), consistent with its reduced GEF2 activity in vitro (23) (Fig. 1F,H).

Distinct Trio variants differentially impact mouse behavior

NDDs affect learning and memory, compulsivity, motor coordination, and social skills, hence we assessed these skills using a diverse array of established behavioral tests (18) in P42 mice bearing Trio variants (Fig. 2A).

Heterozygosity for distinct Trio variants differentially impact NDD-like mouse behaviors.

(A) Schematic illustration of the behavioral tests performed on young adult (P42-P56) heterozygous Trio variant mice of both sexes. All mice proceeded through the same battery of tests. (B) +/K1431M and +/K1918X mice of both sexes had decreased latency to fall off an accelerating rotarod compared to WT male mice. In male mice (left), linear regressions identified differences from WT in slopes, indicating impaired rate of improvement in the skill (WT 16.96 ± 1.344; +/K1431M 7.270 ± 2.019, p<0.0001; +/K1918X 10.61 ± 1.444, p<0.0001; ####p<0.0001) (n=40 WT; 10 +/K1431M; 16 +/K1918X; 13 +/M2145T male mice). In female mice (right), linear regressions identified differences from WT in slopes (+/K1431M 9.436 ± 2.146, p=0.0215; vs WT 14.52 ±1.792; #p<0.05) and intercepts (+/K1918X 6.492 ± 5.555, p=0.0248; vs WT 19.28 ± 5.942; €p<0.05) (n= 28 WT; 11 +/K1431M; 16 13 +/K1918X; and 15 +/M2145T female mice). (C) +/K1431M mice of both sexes and +/K1918X females showed impaired social interactions in a three-chamber test, showing no preference to the (Str.) vs. inanimate object (Obj.) compared to WT. (D) +/K1918X mice of both sexes and +/M2145T females exhibit impaired novel object recognition and spend equal time exploring a novel object (N) and a familiar object (F). (E) Male +/K1918X mice exhibited increased nestlet shredding over 30 min (26.26 ± 3.61% shredded vs WT 14.26 ± 2.97%; p=0.0433), and +/K1431M mice exhibited a trend toward increased nestlet shredding (25.90 ± 4.34% shredded, p=0.1038) compared to WT mice. n=19 male, 19 female WT; 10 male, 10 female +/K1431M; 15 male, 11 female +/K1918X; 9 male, 10 female +/M2145T mice. All data are shown as mean ± SEM, significant differences identified using two-way ANOVA with post-hoc Bonferroni MC (nsp<0.1, *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001). Numbers of mice quantified per group are annotated inside the bar unless otherwise indicated.

We measured motor coordination and learning using an accelerating rotarod test. +/K1431M and +/K1918X mice of both sexes fell from an accelerating rotarod with reduced latency relative to WT littermates, while +/M2145T mice performed similarly to WT (Fig. 2B). In addition, +/K1431M and +/K1918X males and +/K1431M females showed a lower rate of improvement in this skill over repeated trials (Fig. 2B). No deficits in muscle strength were noted in any genotype using the Kondziela inverted screen test prior to rotarod testing.

We used an open field test (OFT) and the elevated plus maze (EPM) to assess general locomotor activity and exploratory behavior. In the OFT (Supp. Fig. 2A-D), Trio variant mice did not differ significantly from WT in total distance traveled or mean speed; however, +/K1431M females spent a greater amount of time in the outer and corner zones of the OFT compared to WT (Supp. Fig. 2D, right), suggesting increased anxiety-like behavior in +/K1431M female mice. WT mice prefer the protected closed arms over the open arms in the EPM (Supp. Fig. 2E,F), whereas +/K1431M and +/K1918X mice of both sexes and +/M2145T females did not exhibit this preference (Supp. Fig. 2F).

Deficits in social communication and social interaction are defining features of individuals diagnosed with NDDs (1). In a three-chamber social interaction test, +/K1431M mice of both sexes and +/K1918X females showed no preference for the stranger mouse over the object (Fig. 2C), while +/M2145T mice of both sexes and +/K1918X males exhibited preference for the stranger mouse, similar to WT (Fig. 2C). In a novel object recognition test, +/K1918X mice of both sexes and +/M2145T females failed to discriminate between the novel and familiar objects, while +/K1431M mice and +/M2145T males exhibited normal discrimination between novel and familiar objects similar to WT (Fig. 2D).

Repetitive behaviors and stereotypies are often identified in individuals with NDDs (64). We found a significant increase in compulsive nestled shredding in +/K1918X and +/K1431M males, with no significant changes in females (Fig. 2E).

Behavioral phenotypes mostly overlapped between +/K1431M and +/K1918X mice, with differences in social interaction and memory tests. +/M2145T mice exhibited the fewest phenotypes in measured tasks. Additionally, some behavioral manifestations appeared to be selective to one sex. Together, these data indicate that these Trio alleles differentially impact behavior.

Trio +/K1431M and +/K1918X mice have smaller brains

TRIO variants that reduce TRIO GEF1 activity are associated with microcephaly (24, 25, 32), so we assessed head and brain size in Trio variant mice (Fig. 3A-E (males), Supp. Fig. 3A-E (females)). After adjusting for body weight, both the head width (Fig. 3D, Supp. Fig. 3D) and brain weight (Fig. 3E, Supp. Fig. 3E) were reduced in P42 adult +/K1431M mice of both sexes relative to WT, consistent with the microcephaly seen in patients harboring TRIO GEF1-deficient alleles. Of note, these ratios were driven by the increased body weight exhibited by P42 +/K1431M mice of both sexes compared to WT counterparts (Fig. 3C, Supp. Fig. 3C). +/K1918X mice of both sexes displayed a reduced brain weight-to-body weight ratio (Fig. 3E, Supp. Fig. 3E) but an unchanged head width-to-body weight ratio (Fig. 3D, Supp. Fig 3D) relative to WT, suggesting a disproportionate reduction in brain mass in +/K1918X mice. Meanwhile, +/M2145T mice did not differ from WT in either head width or brain weight when adjusted for body weight.

Trio +/K1431M and +/K1918X mice have smaller brain weights, but only +/K1918X brains have smaller less complex neurons.

(A) Ear-to-ear head width is reduced in P42 +/K1918X and +/M2145T compared to WT male mice (+/K1431M: 12.40 ± 0.04 mm, p=0.95; +/K1918X: 12.15 ± 0.08, p=0.001; +/M2145T 12.01 ± 0.15, p<0.0001; vs WT 12.49 ± 0.04 mm, n=17-45). (B) Brain weight is significantly decreased relative to WT in P42 males of all three heterozygous Trio variants (+/K1431M: 0.382 ± 0.004 g, p=0.04; +/K1918X: 0.346 ± 0.004 g, p<0.0001; +/M2145T 0.378 ± 0.005 g, p=0.002; vs WT 0.396 ± 0.004 g, n=44-98). (C) Body weight is significantly increased in P42 +/K1431M males and decreased in +/K1918X males (+/K1431M: 22.91 ± 0.38 g, p=0.01; +/K1918X: 20.67 ± 0.03 g, p=0.001; +/M2145T: 21.22 ± 0.33 g, p=0.44; vs WT 21.76 ± 0.19 g, n=45-118). (D) Head widths normalized to body weight of P42 +/K1431M male mice were reduced 10.8% compared to WT mice (+/K1431M: 0.520 ± 0.008 mm/g, p=0.0001; +/K1918X: 0.598 ± 0.012 mm/g, p>0.999; +/M2145T 0.607 ± 0.023 mm/g, p=0.54; vs WT 21.76 ± 0.19 mm/g, n=17-46). (E) Brain weights normalized to body weight of P42 +/K1431M and +/K1918X male mice were reduced 3.9% and 7.9%, respectively compared to WT mice (+/K1431M: 0.520 ± 0.008 mm/g, p=0.0001; +/K1918X: 0.598 ± 0.012 mm/g, p>0.999; +/M2145T 0.607 ± 0.023 mm/g, p=0.54; vs WT 21.76 ± 0.19 mm/g, n=17-46). (F) Representative images of Nissl-stained 30 μm coronal slices of male P42 WT and heterozygous Trio variant brains. (G) Total cross-sectional tissue area of Nissl-stained coronal sections were reduced ∼9% +/K1918X in P42 male mice compared to WT. (H) Representative images of Nissl-stained cortical layers (L1-L6, dotted black box) of P42 WT and heterozygous Trio variant brains. (I) The total cortical thickness (from H) is reduced by ∼8% in +/K1918X P42 male brains compared to WT. (J) Thickness of individual cortical layers, as identified in Nissl stains in Fig. 3H. L2/3 and L5 were preferentially reduced (-12% and -13%, resp.) in +/K1918X cortex relative to WT (L2/3: 0.306 ±0.011 mm vs WT 0.346 ±0.010 mm, p=0.0043; L5: 0.274 ±0.008 mm vs WT 0.314 ±0.008 mm, p=0.0054). (K) Representative traces of M1 L5 PNs from heterozygous male Trio variant mice crossed with Thy1-GFP(M). (L) +/K1918X M1 L5 PNs show a trend toward reduced basal dendritic field size (0.1172 ± 0.0078 mm²; vs WT 0.1368 ± 0.0077 mm², p=0.0933; n=15-22 neurons per mouse), as measured by convex hull analysis of dendrite arbor reconstructions. (M) Both +/K1918X and +/M2145T exhibit significantly smaller apical dendritic field size (+/K1918X: 0.5157 ± 0.0169 mm², p=0.0460; +/M2145T: 0.4893 ± 0.0285 mm², p=0.0062) compared to WT (0.6081 ± 0.0319 mm²; n=15-22 neurons per mouse). All data shown as mean ± SEM. One-way ANOVA with post-hoc Bonferroni MC test identified significant differences from WT (nsp<0.1, *p<0.05, **p<0.01). (N,O) Sholl analysis revealed basal (N) and apical (O) dendritic arborization changes in Trio variant M1 L5 PNs compared to WT: both basal and apical arborization was reduced in +/K1918X, while proximal basal arborization was increased in +/K1431M. Two-way ANOVA (stacked) with post-hoc Bonferroni MC test identified differences from WT.

Trio variant mice show mild changes in cortical organization

We examined whether the decreased brain weight and head size observed in +/K1431M and +/K1918X mice were associated with anatomical defects. Gross histological analyses of fixed Nissl-stained brain sections in P42 male mice showed only mild morphological defects (Fig. 3F-J). We observed reductions in both total cross-sectional brain area (Fig. 3F-G) and cortical thickness (Fig. 3H-I) in +/K1918X brains, consistent with their smaller head size and brain weight. Significant decreases in cortical layer 2/3 and layer 5 thickness were observed in +/K1918X brains (Fig. 3H-J), and were in proportion to the relative decrease in cortical thickness relative to WT (Supp. Fig. 3F). Meanwhile, no significant differences in brain area or cortical thickness were observed in either +/K1431M or +/M2145T mice.

Deletion of Trio has been shown to impair the migration of forebrain interneuron progenitors, resulting in fewer neurons entering the cerebral cortex and altered distribution of cortical layers (19, 65). However, we observed no change in total cortical cell density or in layer-specific cell density in heterozygous Trio variant male mice, nor were numbers of cortical NeuN+ neuronal cells or PV+ inhibitory neurons altered relative to WT, although there were trends toward increased DAPI+ cell density in +/K1918X and increased NeuN+ cell density in +/M2145T motor cortex (Supp. Fig. 3H-N). These data suggest that the reduced brain size of +/K1918X mice results from a loss of neuropil rather than reductions in cortical neuron number.

Trio variant heterozygotes exhibit alterations in dendritic arbors and synaptic ultrastructure

Altered dendritic arbor morphology and dendritic spine abnormalities are hallmarks of NDDs (66–72). Excitatory neuron-specific ablation of one or both Trio alleles decreased dendritic arborization, increased spine density, and yielded smaller synapses in cortex area M1 Layer 5 pyramidal neurons (M1 L5 PNs) (18).

Sholl analysis of M1 L5 PNs revealed significant reductions in both basal and apical dendritic arbor complexity and dendritic field size in P42 +/K1918X neurons, an increase in proximal basal arbor complexity in +/K1431M neurons, and a decrease in distal apical arbor complexity and apical dendritic field size in +/M2145T neurons relative to WT (Fig. 3K-O, Supp Fig. 3.G)

Conditional Trio knockout specifically in forebrain excitatory neurons significantly increased the dendritic spine density on M1 L5 PNs, although they were smaller and appeared more immature (18). Notably, none of the Trio variant heterozygotes exhibited altered dendritic spine density on M1 L5 pyramidal neurons compared to WT mice on either apical or basal secondary arbors (Supp. Fig. 4L,M).

Electron microscopy of cortical area M1 L5 revealed that synapse density was significantly increased in +/K1918X mice compared to WT (Fig. 4A,B), possibly due to a net reduction in neuropil resulting from smaller dendritic arbors. Within synapses, postsynaptic density (PSD) length was slightly decreased in +/K1918X (by 6%) and +/M2145T (by 6.6%) mice relative to WT, but not in +/K1431M mice (Fig. 4C). Cross-sectional presynaptic bouton area and spine head area were similar to WT in each Trio variant heterozygote (Fig. 4D,E).

Trio variants differentially impact synapse ultrastructure and synaptic vesicle distribution.

(A) Representative electron micrographs (EMs) from motor cortex layer 5 (M1 L5) of P42 WT and Trio variant mice. Post-synaptic regions are pseudo-colored in cyan; pre-synaptic regions in magenta. (B) Asymmetric synapse density was increased in +/K1918X mice (0.09205 ± 0.004775 synapses/um2; vs WT 0.07633 ± 0.003954 synapses/um2, p=0.0345). (C) PSD lengths were slightly decreased in M1 L5 synapses of by 6% in +/K1918X and 6.6% in +/M2145T mice vs WT (+/K1918X 0.2926 ± 0.004652 um, p=0.0204; +/M2145T 0.2916 ± 0.004922 um, p=0.0142; vs WT 0.3125 ± 0.005612 um). (D,E) Presynaptic bouton and spine head areas of Trio variants M1 L5 synapses were unchanged from WT. (F) Synaptic vesicles (SVs) distribution per 100 nm of active zone (AZ) length in M1 L5 as a function of distance from the AZ. +/M2145T showed an increase in readily releasable pool (RRP) identified as docked SVs (15 nm from AZ; 1.23 ± 0.05 vs WT 0.90 ± 0.05) and increase in tethered SVs (50 nm from AZ; 1.44 ± 0.04 vs WT 1.20 ± 0.05). +/K1918X and +/M2145T also showed an increase in the reserve pool of SVs (200 nm from AZ; 3.51 ± 0.21 and 3.81 ± 0.18, resp. vs WT 2.74 ± 0.16, n=15-30 synapses/mouse). (G) Total releasable pool, calculated as number of SVs at 15-150 nm from AZ per area of distribution (nm2). RRP (15-50 nm from AZ) was significantly increased in +/M2145T (0.257 ± 0.007 vs WT 0.228 ± 0.008), driven by increased docked and tethered SVs. All data are presented as mean ± SEM, significance tested by ordinary one-way ANOVA with post-hoc Bonferroni MC test (*p<0.05, **p<0.01, ***p<0.001, ****p<0.0001).

Synaptic vesicle (SV) distribution was significantly altered in +/M2145T mice relative to the other genotypes (Fig. 4F,G). The densities of both docked and tethered vesicles at 15 and 50 nm from the active zone (AZ) respectively, which together can estimate the readily releasable pool (RRP) (52, 53), were significantly increased in +/M2145T relative to the other genotypes (Fig. 4F,G). Also, SV distribution at 200 nm from the AZ, thought to contribute the reserve pool, were also significantly increased in +/M2145T mice, suggesting an overall larger SV pool size. No differences in synaptic vesicle distribution were noted in+/K1431M mice, while +/K1918X showed modestly increased SVs at 200 nm from AZ relative to WT.

Synaptic transmission and plasticity are differentially impaired by distinct Trio variants

Loss of Trio function or disruption of TRIO GEF1 activity in slice culture decreases AMPAR levels at excitatory synapses (18, 20, 21, 28), while mice bearing a GEF1-deficient Trio allele exhibited decreased gamma-aminobutyric acid receptor (GABAR)- and glycine receptor (GlyR)-mediated inhibitory miniature current (mIPSC) frequencies in the prefrontal cortex (19). To explore how Trio variants impact synaptic function, we measured both miniature excitatory currents (mEPSCs) and mIPSCs in M1 L5 PNs in each of the Trio variant heterozygotes at P35-42 consistent with other experiments.

AMPAR-mediated mEPSC amplitudes were significantly increased in +/K1431M and +/K1918X mice relative to WT littermates, with no change in their frequencies (Fig. 5A-C). In contrast, AMPAR-mediated mEPSC amplitudes were unchanged in +/M2145T mice, but their frequencies were significantly increased. No significant changes in NMDAR mEPSC amplitudes were noted, while NMDAR mEPSC frequencies were decreased in +/K1431M and increased in +/M2145T mice cortex (Fig. 5D-F). Notably, these findings correlated with gross alterations in the ratio of NMDAR/AMPAR-mediated evoked (e)EPSCs measured in M1 L5 PNs following stimulation in L2/3. +/K1431M and +/K1918X Trio variant heterozygotes showed decreased NMDA/AMPA ratios, indicating imbalances in NMDAR-versus AMPAR-mediated conductance (Fig. 5J,K) and suggesting an increase in synaptic AMPAR signaling in both mice. Significant decreases in mIPSC frequencies were noted in +/K1431M and +/M2145T mice relative to WT mice, with no change in amplitudes (Fig. 5G-I). +/K1918X mice exhibited increased mIPSC amplitude with no observed change in mIPSC frequency. Together, these data indicate that the Trio variants differentially impact excitatory and inhibitory transmission.

Trio variant mice exhibit deficits in synaptic signaling and LTP.

(A,D) Representative traces of miniature excitatory AMPAR-mediated mEPSCs, NMDAR-mediated mEPSCs, and (G) inhibitory postsynaptic currents (mIPSCs) in M1 L5 pyramidal neurons of WT and Trio variant mice. (B) AMPAR-mediated mEPSC amplitudes were significantly increased in +/K1431M (16.67 ± 1.04 pA; p=0.0009) and +/K1918X (14.71 ± 0.92 pA; p=0.03) slices, with no observed changes in +/M2145T slices (13.90 ± 1.16 pA; p=0.16) compared to WT (11.25 ± 0.84 pA; n=17-25 neurons from ≥6-8 mice per group). (C) No significant changes in AMPAR mEPSC frequencies (q) were observed in +/K1431M and +/K1918X, while +/M2145T had an increase (2.20 ± 0.15 1/s; vs WT 1.55 ± 0.09 1/s; p=0.0005). (E, F) NMDAR mEPSC frequencies were reduced in +/K1431M (0.89 ± 0.12 1/s; vs WT 1.3324 ± 0.11 1/s; p=0.015) and showed an increase in +/M2145T mice (1.68 ± 0.10 1/s vs WT 1.3324 ± 0.11 1/s; p=0.044, n=9-13 neurons from ≥5-7 mice per group). (H, I) GABA/GlyR mIPSC amplitudes were significantly increased in +/K1918X vs WT (23.69 ± 2.89 pA; vs 15.86 ± 1.56 pA, respectably; p=0.008), while frequency was decreased in +/K1431M and +/M2145T (0.94 ± 0.14 1/s, p<0.0001; and 1.64 ± 0.19 1/s, p=0.013; respectably; vs WT 2.44 ± 0.20; n=16-26 neurons from ≥6-8 mice per group). (J) Representative averaged traces of NMDA and AMPA eEPSCs recorded in M1 L5 PNs. (K) Heterozygous +/K1431M and +/K1918X Trio variants mice display reduced NMDAR/AMPAR eEPSC amplitude ratios, suggesting an increase in AMPAR transmission in M1 L5 PNs (+/K1431M: 0.75 ± 0.06, p=0.0002; +/K1918X: 0.69 ± 0.05, p<0.0001; +/M2145T: 1.00 ± 0.08, p=0.37; vs WT: 1.15 ± 0.07; n=13-19 neurons from ≥5-6 mice per group). (L) Averaged representative traces of baseline and post-TBS eEPSC currents in M1 L5 PNs of WT and Trio variant mice. (M) Normalized eEPSC amplitudes measuring LTP in L5 PNs by TBS in L2/3 afferents in all genotypes showed a significant decrease in the initiation and no potentiation of the LTP in +/K1431M and +/K1918X, with increase in initiation and potentiation of +/M2145T M1 L5 PNs compared to WT. LTP was induced at 0 min. RM two-way ANOVA with post-hoc Bonferroni MC test identified significant differences (n=6-8 neurons from ≥4-5 mice per group). Data are presented as mean ± SEM; significance tested by one-way ANOVA with post-hoc Bonferroni test unless specified otherwise (**p<0.01; ***p<0.001; ****p<0.0001).

Finally, we tested the ability of the L5 PNs to undergo long-term potentiation (LTP) following theta-burst stimulation of L2/3 afferents (Fig. 5L,M). While LTP was robustly induced and potentiated in M1 L5 PNs from WT mice, LTP induction and potentiation were deficient in slices from +/K1918X and +/K1431M mutant mice. In contrast, +/M2145T L5 PNs showed increased induction and prolonged potentiation of LTP compared to WT L5 PNs.

Neurotransmitter release is altered in Trio +/K1431M and +/M2145T heterozygotes

In addition to its postsynaptic roles, Trio localizes presynaptically and interacts with the presynaptic active zone scaffolding proteins Bassoon and Piccolo (13, 14, 21). Additionally, recent work demonstrates that Rac1 activity levels can bidirectionally affect SV probability of release (Pr) in excitatory synapses (14, 56, 80). To assess the impact of Trio variants on presynaptic function, we first measured the paired-pulse ratio (PPR) to test possible changes in Pr. Synapses with a low probability of release (Pr), such as L2/3 onto L5 PNs synapses, exhibit paired-pulse facilitation (PPF), where synaptic response increases for the second of two apposed stimuli due to elevated residual Ca²⁺ promoting SV fusion. WT and +/K1918X M1 L5 PNs exhibited normal facilitation of eEPSC amplitudes that decreased with increased interstimulus interval (ISI) (Fig. 6A,B). PPF was significantly enhanced in +/K1431M M1 L5 PNs at short ISIs relative to WT, suggesting a reduction in Pr at these synapses (Fig. 6A,B). The +/M2145T PPR curve was complex, with significantly reduced PPF at short ISIs, yet clearly increased PPF at longer ISIs compared to WT (Fig. 6A,B). The decreased PPF at initial ISIs in +/M2145T mice, together with an increase in both AMPAR and NMDAR mEPSC frequency (Fig. 5A-C), demonstrate an increase in both synchronous and spontaneous glutamate Pr.

Trio +/K1431M and +/M2145T variant mice have deficiencies in short-term facilitation, glutamate Pr and RRP.

(A) Representative traces in M1 L5 PNs of WT, Trio variant mice in response to paired pulse stimulation in L2/3. (B) Paired-pulse ratio (PPR) at varying interstimulus intervals (ISIs) were overlaid with a single exponential fit (except for +/M2145T data). An increase in the initial PPR was observed in M1 L5 PNs of +/K1431M slices (35 ms: 1.70 ± 0.089, p=0.003; 60 ms:1.40 ± 0.07, p=0.046; 100 ms: 1.27 ± 0.05, p=0.031; n=20-34 neurons from ≥7-9 mice per group) with no change in +/K1918X slices; and in +/M2145T slices we observed a decrease in initial RRP at shorter ISIs (35 ms: 1.05 ± 0.06, p<0.0001; 60 ms: 0.97 ± 0.06, p=0.037) and an increase at longer ISIs (100 ms: 1.36 ± 0.09, p=0.034; 200 ms: 1.18 ± 0.08, p=0.013) compared to WT (35 ms: 1.40 ± 0.04; 60ms: 1.21 ± 0.03; 100 ms: 1.13 ± 0.03; 200 ms 1.0 ± 0.02; 300 ms 0.96 ± 0.17). (C) Representative traces of AMPAR eEPSCs in M1 L5 PNs under HFS (15 pulses at 40 Hz) in L2/3. (D) AMPAR eEPSCn amplitudes normalized to eEPSC1 of the train revealed changes in the depletion rates during HFS in Trio +/K1431M and +/M2145T variants compared to WT (tau decay (Td), WT: 2.7 s, +/K1431M: 3.19 s, +/M2145T: 4.79 s, +/K1918X: 2.52 s; n=12-15 neurons from 5-7 mice). (E) The estimated glutamate probability of release (Pr) was decreased in +/K1431M slices (0.13 ± 0.099; p=0.013) and increased in +/M2145T slices (0.26 ± 0.019, p=0.042), with no significant change in +/K1918X slices (0.15 ±0.01, p=0.64) compared to WT M1 L5 PNs (0.19 ± 0.01; n=12-15 neurons from ≥ 5 mice per group). (F) The calculated size of the readily releasable vesicle pool (RRP) was increased only in +/M2145T M1 L5 PNs compared to WT (665.7 ± 68.5 pA vs 415.8 ± 43.9 pA, p=0.012). RRP in +/K1431M and +/K1918X synapses did not differ from WT (543.1 ± 64.4 pA; and 543.1 ± 64.4 pA, respectively vs 415.8 ± 43.9 pA) (G) Exponential fits of the fractional recovery plotted vs ISI, to estimate synapse ability to recover from RRP depletion. Time of recovery, measured by exponential tau recovery (TR), was significantly decreased in +/K1431M M1 L5 PNs (5.7 s, vs WT 2.2 s), +/K1431M also exhibited an inability to fully recover to initial levels after ISI 18 s, vs WT. Data are presented as mean ± SEM, with significant differences from WT tested using one-way ANOVA with post-hoc Bonferroni (*p<0.05; **p<0.01; ***p<0.001; ****p<0.0001).

To further characterize changes in neurotransmitter release in these mice we used high frequency (40Hz) stimulation (HFS) trains to quantitatively estimate glutamate Pr, RRP size, and rates of SV depletion and recovery. A plot of the normalized eEPSC responses to a HFS train stimulation again revealed facilitation upon the first 2-3 stimulations that was increased in +/K1431M and decreased in +/M2145T slices relative to WT, with no changes in +/K1918X slices (Fig. 6C,D). Initial facilitation was followed by decaying eEPSC amplitudes, reflecting SV depletion under HFS. +/K1431M and +/M2145T Trio variants exhibited a slower train decay rate relative to WT during HFS with +/M2145T depleting at half the rate of WT (Td, +/K1431M: 3.19 s, +/M2145T: 4.79 s vs WT: 2.70 s) (Fig. 6D).

We used a ‘Decay’ method (54, 55) to estimate Pr and RRP size from HFS trains which allows us to account for initial facilitation seen in the train eEPSCs. Glutamate Pr in the L2/3-L5 synapses was increased in +/M2145T mice, while it was decreased for +/K1431M mice (Fig. 6E), consistent with the relative changes observed in PPF for these mice. RRP size was much larger in L2/3-L5 synapses of +/M2145T mice relative to WT (Fig. 6F), consistent with the increased SV distribution found at 15-50 nm from the AZ in electron micrographs (Fig. 4F,G). We tested the ability of Trio variant heterozygotes to recover after train depletion by pairing HFS train with a single stimulus at increasing intervals (0.01, 2, 6, 9, 12, 18s) and calculating the fractional recovery (see Methods). The recovery rate (TR) was significantly slower in +/K1431M L5 PNs and they did not recover to their initial strength within 18s, plateauing at 78% of maximal recovery compared to WT (Fig. 6G). Together, Trio +/K1431M and +/M2145T mice exhibit distinct and significant alterations in short-term plasticity and synchronous glutamate release.

Trio variant cortex displays different proteomic signatures

We used comparative proteomics from P21 cortex to identify proteins and pathways that were differentially altered by the Trio variants. We quantified a total of 7,362 proteins, finding distinct differences in the cortical proteome for each genotype (Supp. Fig. 5, Supp. Table 1). Gene Set Enrichment Analysis (GSEA) (57, 58) revealed alterations in distinct functions for each Trio variant (Fig. 7A; Figure 7 -Source data 1). Of note, the only enriched gene set specific to neurons was downregulation of the synaptic vesicle pathway in +/K1431M cortex; all other gene sets were not cell type-specific.

Trio variant mice show different molecular changes in the cortex involving presynaptic machinery and Rac1 GEFs.

(A) Bar graph illustrating the top enriched pathways (FDR q-value<0.2, *FDR<0.05) identified by gene set enrichment analysis (GSEA) for each Trio mutant mouse compared to WT, using all 7,362 proteins quantified by mass spectrometry in P21 cortex (n=4/genotype), sorted by normalized enrichment score (NES). Pathways with +NES are upregulated, -NES are downregulated vs. WT. [] indicates gene set: [R] Reactome, [WP] WikiPathways, [K] KEGG. Full list attached in Figure 7 – Source data 1. (B,C) Bar graphs illustrating the top enriched (FDR q-value <0.001) (B) cellular components and (C) biological processes identified by GSEA, using synaptic proteins from SynGO gene sets (n=1077 proteins), full list see in Figure 7 - Source data 2. (D) Representative immunoblots in synaptosomes isolated from P42 cortex of WT and Trio variant mice. (E-H) Normalized intensity levels from immunoblots demonstrate significant increases of (E) Munc18-1 (also known as syntaxin binding protein1), (F) synaptophysin (Syp), (G) syntaxin1a (Stx1) and (H) synaptotagmin3 (Syt3) levels in +/M2145T synaptosomes; Syp is increased while Stx1a is significantly decreased in +/K1431M synaptosomes compared to WT. Ordinary one-way ANOVA with post-hoc Bonferonni MC test identified differences from WT (*p<0.05, **p<0.01, ***p<0.001; n=synaptosomes from 7-14 male mice). (I) Representative immunoblots of select RhoGEFs from P42 cortical lysates of WT and Trio variant mice. (J-L) Normalized intensity levels from immunoblots identified ∼47% increase of Tiam1 levels in +/K1431M and increase ∼45% in +/M2145T cortex vs to WT; VAV2 is increased ∼34% in +/M2145T cortex compared to WT. Unpaired t-tests identified differences from WT (*p<0.05; n=6 mice per genotype).

Figure 7 – Source data 1. Gene Set Enrichment Analysis of P21 cortex proteome of Trio WT, +/K1431M, +/K1918X, and +/M2145T mice.

Figure 7 - Source data 2. Synaptic Gene Ontologies (SynGO) of P21 cortex proteome of Trio WT, +/K1431M, +/K1918X, and +/M2145T mice.

We used SynGo (59) to investigate whether the Trio variants impacted synaptic functions. 1,067 of the 7,362 total quantified proteins were synaptic proteins listed in the SynGO geneset (Figure 7, Figure 7 - Source data 2). When restricted to brain-specific genes, all three Trio variant heterozygotes showed enrichment of differentially expressed proteins (DEPs) in synaptic processes (Fig. 7B,C). Notably, +/M2145T upregulated DEPs and+/K1431M downregulated DEPs were significantly enriched at the presynapse (Fig. 7B), but with only +/K1431M downregulated DEPs showing significant enrichment for postsynaptic receptor and synaptic vesicle cycling (Fig. 7C). Meanwhile, +/K1918X showed enrichment of DEPs at the postsynapse, particularly in the postsynaptic density (Fig. 7B). Together, our proteomics data point to a significant deficit in presynaptic function in both +/K1431M and +/M2145T cortex, as well as a significant effect of +/K1431M on postsynaptic function.

Rho GEFs and synaptic regulatory proteins are altered in Trio +/K1431M and +/M2145T heterozygotes

Given our findings from proteomic analysis and electrophysiology, we measured levels of key presynaptic regulators, including synaptophysin (Syp), syntaxin binding protein1 (Stxbp1, also known as Munc18-1), syntaxin1a (Stx1a) and synaptotagmin3 (Syt3), which are crucial for synaptic vesicle (SV) tethering, docking, replenishment, and calcium-dependent replenishment, respectively. We also measured levels of presynaptic proteins in P42 cortical synaptosomes, where these proteins are enriched. Munc18-1, Syt3, and Syp levels were increased in +/M2145T synaptosomes relative to WT (Fig. 7D-H). Meanwhile, Stx1a levels were significantly decreased in +/K1431M synaptosomes compared to WT, with no significant changes in +/K1918X compared to WT mice.

The elevated Rac1 activity in +/K1431M brain lysates and synaptosomes (Fig. 1I) seemed at odds with previous reports that K1431M reduces TRIO GEF1 activity (Supp. Fig. 1A,B) (19, 23, 28). We hypothesized that homeostatic compensation in +/K1431M mice may alter expression of other RhoGEFs and GAPs. Indeed, levels of the Rac1 GEF Tiam1 were increased in both +/K1431M and +/M2145T P42 cortical lysates, while VAV2 levels were increased in +/M2145T P42 lysates (Fig. 7I-L). Levels of the Trio paralog Kalirin (73) were unaffected in the Trio variant mice at P42 (Fig. 7J). Together, our proteomic analyses suggest that presynaptic functions are altered in +/K1431M and +/M2145T mice and may be driven by abnormal levels of crucial presynaptic regulatory proteins and changes in Rac1 and RhoA activity.

NSC23766, a Rac1-specific inhibitor, rescues neurotransmitter release in Trio +/K1431M heterozygotes

Rac1 negatively regulates synaptic vesicle replenishment and synaptic strength in excitatory synapses (14, 74). In +/K1431M synapses, increased Rac1 activity and decreased Stx1a levels were associated with reduced synaptic strength and impaired vesicle replenishment. We tested if the acute application of the Rac1 inhibitor NSC23766 (NSC) could rescue these deficits. Treatment of +/K1431M and WT slices acutely with NSC shifted the PPF downwards in M1 L5 PNs, suggesting an increase in Pr in both cases (Fig. 8A,B). Notably, at longer ISI, PPRs in NSC-treated WT slices still normalized at around 1, while PPRs in NSC-treated +/K1431M slices shifted below 1, exhibiting slight depression.

NSC, Rac1 inhibitor application rescued Pr in +/K1431M L23-L5 synapses and improves SV recycling.

(A) Representative PPR traces of WT and Trio +/K1431M slices with or without 5 min application of 100 µM NSC23766. (B) Acute application of NSC onto both +/K1431M and WT synapses lead to decrease in PPF in M1 L2/3-L5 synapses. +/K1431M slices significantly shifted the PPF curve at all ISI downwards compared to untreated +/K1431M slices, and showed no significant difference from WT (+/K1431M +NSC 35 ms: 1.25 ± 0.06, p<0.0001; 60 ms: 1.13 ± 0.052, p=0.0007; 100 ms: 1.02 ± 0.053, p=0.0017; 200 ms 0.91 ± 0.039, p=0.0043; 300 ms 0.88 ± 0.045, p=0.021), with +/K1431M shifting into paired pulse depression at 200-300 ms intervals, while WT PPF plateauing to 1. (C) Representative traces of AMPAR eEPSCs in M1 L5 PNs under HFS of WT and Trio +/K1431M slices before and after NSC application. (D) Normalized AMPAR eEPSCn amplitudes of the train revealed changes in the depletion rates during HFS before and after NSC application to WT and +/K1431M slices (tau decay (Td), WT+NSC: 2.85 s vs WT: 2.70 s; +/K1431M+NSC: 2.66 s vs +/K1431M: 3.19 s, n=12-15 neurons from 5-7 mice). (E) Rac1 inhibition by NSC increased the glutamate Pr in both WT and +/K1431M slices (WT + NSC 0.25 ± 0.067 vs initial 0.19 ± 0.01, p=0.046; and for +/K1431M + NSC 0.23 ± 0.019 vs initial 0.13 ± 0.099, p<0.0001; n=15-18 neurons from ≥ 5 mice per group). (F) RRP in WT or +/K1431M synapses with NSC did not show significant changes to initial values (WT +NSC: 370.3 ± 82.37 pA vs 415.8 ± 43.9 pA, pL0.99; +/K1431M + NSC: 427.9 ± 79.2 vs 543.1 ± 64.44 pA, pL0.99). (G) Exponential fits of the fractional recovery for WT and +/K1431M with and without NSC application. NSC application led to a faster recovery time in WT (+NSC: 1.5 s vs initial 2.2 s) and it significantly improved but did not fully rescue recovery time in +/K1431M (+ NSC 3.2 s vs initial 5.7 s), but allowed for full recovery at 18s. Data are presented as mean ± SEM, significance tested using one-way ANOVA with post-hoc Bonferroni (*p<0.05; **p<0.01; ***p<0.001; ****p<0.0001).

Application of NSC to WT and +/K1431M slices under HFS train stimulation decreased initial facilitation in both (Fig. 8C,D), and significantly increased Pr (Fig. 8E) without affecting RRP size (Fig. 8F) in both, rescuing the initial Pr in +/K1431M slices (Fig. 8C-F). NSC treatment also led to a faster train decay rate ([d) in +/K1431M slices but did not significantly change WT relative to untreated slices (+/K1431M +NSC [d, 2.66 s vs. initial 3.19 s; WT +NSC: 2.85 vs. 2.70 s) (Fig. 8D).

Finally, we tested if acute Rac1 inhibition impacts the rate of recovery of the RRP following HFS stimulation. NSC treatment increased the fractional recovery rate (TR) in both WT and +/K1431M slices (WT +NSC: 32% faster than WT; +/K1431M +NSC: 40% faster than +/K1431M). NSC treatment of the +/K1431M variant allowed for full recovery at 18s interval and sped up the recovery rate, but it remained significantly slower compared to WT (by 45%) (Fig. 8G).

Overall, we demonstrate that presynaptic Trio GEF1-dependent Rac1 signaling is crucial for maintaining synchronous glutamate Pr and SV replenishment at cortical L2/3-L5 synapses.

Discussion

Large-scale genetic studies show significant overlap in risk genes for ASD, SCZ, and BPD, many converging on synaptic proteins (75–83). However, how variants in a single gene contribute to different NDDs remains a major unresolved question. Our study reveals that mice heterozygous for NDD-associated Trio variants differentially affecting Trio protein levels or GEF activity yield overlapping but distinct behavioral, neuroanatomical, and synaptic phenotypes. Our findings extend prior work demonstrating that Trio is critical for postsynaptic signaling and synaptic plasticity. We also demonstrate for the first time in mice that Trio is critical for glutamate release and synaptic vesicle recycling, and that NDD-associated variants differentially impact these pre- and post-synaptic roles.

Heterozygosity for Trio variants in mice yields phenotypes similar to those observed in NDDs

Individuals with mutations in TRIO present with a range of neurodevelopmental disorder-associated clinical features, including varying degrees of intellectual disability, altered head size, skeletal and facial features, and behavioral abnormalities (22, 24–27, 32, 84). Patients with missense or truncating variants in TRIO that reduce GEF1 activity have mild developmental delay and microcephaly (24, 25, 27, 32). Similarly, we found that heterozygosity for the GEF1-deficient K1431M missense or the K1918X nonsense variants significantly reduced brain weight and/or head size compared to WT mice, along with multiple behavioral impairments. Notably, while both showed impaired motor coordination and learning, only mice bearing the ASD-associated K1431M allele exhibited social interaction deficits. In addition, we observed behavioral differences in male versus female Trio variant mice, possibly similar to human sex differences in the susceptibility to and clinical presentation of NDDs (85).

Both +/K1431M and +/K1918X adult mice of both sexes had reduced brain-to-body weight ratios compared to WT, but these were driven by different factors. The smaller brain size in +/K1918X male mice was associated with a reduction in neuropil and reduced cortical thickness, similar to mice bearing excitatory neuron-specific ablation of one Trio allele (18) and paralleling the reduced gray matter volume and cortical thickness in SCZ patients (86–90). Meanwhile, +/K1431M mice of both sexes exhibited an overall increase in body weight leading to relatively decreased head width- and brain-to-body weight ratios in these mice. Adult +/K1431M male mice had no change in neuropil or cross-sectional brain area, consistent with a prior study describing normal brain size in +/K1431M at E14.5 (19). Rac1 mediates glucose-stimulated insulin secretion from pancreatic islet beta-cells (91–96), which may explain how chronic alterations in Rac1 activity contribute to weight changes in Trio +/K1431M and +/K1918X mice. Most studies of Trio variants have focused on neuronal effects, but expression of Trio in other tissues could explain the increased body weight in these mice, as well as the musculoskeletal abnormalities associated with TRIO variation in humans (22, 24, 26, 31).

Trio variants differentially impact dendritic arbor structure

Rac1 and RhoA signaling is critical for dendrite development (9–12, 97). We found relatively subtle effects of Rac1/RhoA-altering Trio variants on cortical L5 PN dendrites. +/K1918X L5 pyramidal neurons are smaller and less complex than WT neurons, especially in the basal compartment corresponding to L5 where EM images were obtained, consistent with the smaller brain size and reduced cortical thickness of +/K1918X mice. We posit that due to their smaller dendritic field size, L5 neurons pack more densely, contributing to the mildly increased synapse density observed in +/K1918X M1 L5 cortex. Consistent with this hypothesis, we observed a trend toward increased DAPI+ cell density in M1 L5 of +/K1918X neurons.

The reductions in dendritic arborization and length in +/K1918X neurons are consistent with reports of reduced gray matter volume and dendrite alterations in individuals with schizophrenia versus controls (98, 99), but were modest compared to NEX-Trio+/fl mice lacking one copy of Trio in excitatory neurons (18). Reduced Trio function in other cell types, such as in inhibitory neurons or glia, or in neurons from other brain regions that project to cortical M1 L5 PNs, may ameliorate the phenotypes in excitatory neurons of +/K1918X relative to NEX-Trio+/fl mice. In addition, changes to L5 PN dendrites in Trio variant mice appeared to be regionally selective within the arbor: +/K1431M neurons had increased arborization only in proximal basal dendrites, while +/M2145T neurons had decreased arborization only in the most distal apical dendrites. These differences may reflect the differential spatiotemporal influence of Trio in regulating Rac1 versus RhoA activity. Alternatively, given our finding for presynaptic roles for Trio, these differences may reflect differential effects of the Trio variants on both excitatory and inhibitory afferent synaptic inputs which play critical roles in shaping the apical and basal dendrites of L5 PNs (100).

Trio variants impact brain Rho GTPase signaling

We show here that the K1431M variant significantly reduces TRIO GEF1 nucleotide exchange on Rac1 in vitro, consistent with previous reports (23, 28). Decreased Rac1 activity was observed in +/K1431M mice at embryonic day 14.5 in the ganglionic eminence, which is enriched for pre-migratory GABAergic interneurons (19). In contrast, we show an increase in active Rac1 levels in +/K1431M postnatal brains (at P0, P21, P42) and synaptosomes (at P42) from cortex, primarily composed of differentiated excitatory neurons. This increase aligns with observed phenotypes (e.g. reduced Pr) and reveals why reductions in brain volume, dendrites and spines, or AMPAR signaling anticipated from reduced Rac1 activity (101–105) were not observed in +/K1431M mice. We propose that the increased Rac1 activity we observed reflects homeostatic compensation in Rac1 regulation occurring between birth to adult ages, and identify changes in Rac1-specific GEFs, e.g. Tiam1 and Vav2, that may contribute to this compensation. Importantly, we did not find changes in Kalirin levels in the adult brain of these Trio mutant mice, suggesting that Kalirin does not compensate for loss of Trio GEF1 activity at this age.

We observed a significant reduction in active RhoA only in purified synaptosomes of +/M2145T brains, reflecting the synaptic compartment as a key locus of Trio function. In addition, despite +/K1918X mice having half the WT levels of Trio protein, we measured little to no change in active Rac1 and RhoA levels in +/K1918X brains. These findings are consistent with recent evidence that the spatiotemporally precise balance of Rac/Rho activity rather than absolute activity levels can mediate cytoskeletal rearrangements (6, 8, 106, 107). Alterations in the activity of additional potential TRIO substrates, such as RhoG (108), Cdc42 (5), and the neurodevelopmentally critical Rac3 (109–111) could also contribute phenotypes in these mice.

Our results strongly complement a growing body of evidence showing that altered activation of Rho GTPases is an important mechanism affecting synaptic function in ASD (85, 112–116). This highlights Rho GTPase signaling as a convergent pathway in ASD and therefore an attractive target for pharmacotherapy for these disorders.

The NDD-associated Trio variants cause synaptic transmission, plasticity and excitatory/inhibitory imbalance

Overexpression of a TRIO K1431M variant with reduced GEF1 activity decreased AMPAR-mediated mESPC amplitudes in rat organotypic slices (20), while Rac1 activation increased AMPAR amplitudes by promoting synaptic AMPAR clustering (105, 117). In +/K1431M mice, we observed a decrease in the NMDA/AMPA ratio and NMDA mEPSCs frequency, with increased AMPAR mEPSCs amplitudes, with no changes in dendritic spines suggesting a possible reduction in silent synapses due to increased AMPAR incorporation. This increase in AMPAR was associated with increased active Rac1 levels measured in synaptosomes in adult +/K1431M mice. AMPAR mEPSC amplitudes were also increased in L5 PNs of +/K1918X mice, corresponding with a decrease in the NMDA/AMPAR ratio, but no changes in mEPSCs frequencies, also suggesting an increased number of synaptic AMPARs. In +/K1918X cortex, Rac1 activity is reduced at P0 but increases by P21 and P42 to WT levels, suggesting that a relative increase in Rac1 during development may enhance AMPAR tone and affect LTP, especially given the precise spatiotemporal requirements of Rac/Rho activity and their effectors for LTP (105, 117–120). The +/M2145T variant showed no changes in AMPAR or NMDAR amplitudes but had increased mEPSC frequencies in both, without changes in spine or synapse density, consistent with enhanced spontaneous neurotransmitter release.

mIPSC frequencies were decreased in +/K1431M and +/M2145T L5 PNs, while mIPSC amplitude was increased in +/K1918X slices. Sun et al. noted a similar deficit in inhibitory function in +/K1431M prefrontal cortex correlated with reduced interneuron migration to this region, including parvalbumin positive (PV+) neurons (19). In contrast, we did not observe reduced PV+ interneuron numbers in the motor cortex in any Trio variant mice, suggesting that Trio variants may also impact the number of inhibitory synapses or transmission. However, we cannot fully exclude potential defects in migration of other interneuron populations as contributors to these phenotypes.

Overall, heterozygosity for Trio variants dysregulates excitatory and inhibitory synaptic transmission in different patterns, resulting in E/I imbalance, a known driver of NDD phenotypes.

Trio-deficient excitatory neurons are unable to undergo long-term potentiation (LTP) in mouse brain slices (18), which is crucial for working memory in mammals. During LTP, Rac1 is suggested to be transiently activated and deactivated to regulate AMPAR (105, 117–120). Both +/K1431M and +/K1918X L5 PNs exhibited reduced LTP induction and maintenance. The increased AMPAR resulting from elevated Rac1 activity (+/K1431M mice) or the inability of reduced levels of Trio to activate Rac1 (+/K1918X mice) may preclude LTP in these mice.

+/M2145T mice showed a striking increase in the induction and maintenance of LTP, correlating with an increased glutamate Pr and SV pool size. The function of RhoA in plasticity is unknown, but the decrease in RhoA activity and increased levels in presynaptic machinery proteins measured in cortical synaptosomes may underlie the increase in LTP in these mice.

Trio GEF1 and GEF2-deficient variants lead to opposing defects in the synaptic release of glutamate

In neuroendocrine, pancreatic beta, and mast cells, Trio GEF1 activity, Rac1 and RhoA are required for regulation of exocytosis (121–125). Recent work suggests that Trio GEF1 can act through Rac1 to regulate presynaptic processes – Rac1 colocalizes with SVs in the axonal boutons to negatively regulate action potential-dependent (synchronous) glutamate Pr and SV replenishment (14, 74, 126, 127). Using the Rac1 inhibitor NSC, we demonstrate that elevated Rac1 activity, potentially driven by Tiam1 upregulation, drives the reduction in synchronous glutamate Pr and SV replenishment without affecting RRP in +/K1431M mice. NSC application rescues Pr and enhances SV replenishment in +/K1431M slices while also increasing both processes in WT. Decreased Stx1 levels in +/K1431M synapses, a key component of the priming machinery, may still be a limiting factor in effective priming, thus preventing full rescue of the SV recovery by NSC. Overall, NSC had a significant yet weaker effect in WT than in +/K1431M slices, likely due to the overall increased Rac1 activity in +/K1431M slices providing more of a target for NSC.

Conditional knock out of Rac1 in the Calyx of Held increased spontaneous (action potential-independent) glutamate Pr (74), aligning with the hypothesis that Rac1 activity negatively regulates neurotransmitter release by increasing the assembly of actin filaments at the AZ that impede SV fusion. +/K1431M cortical slices showed significant increases in AMPAR amplitudes, and our data are consistent with fewer silent synapses. Yet, we observed no change in AMPA mESPC frequency. We hypothesize that the Rac1-dependent decrease in Pr identified in these slices decreased AMPAR mEPSCs frequency; hence, the opposing changes from these pre- and post-synaptic effects mask the expected change in AMPAR mEPSC frequency.

RhoA activity is significantly reduced in cortical synaptosomes of +/M2145T mice, and this is associated with increased RRP size. While a specific function for RhoA in regulating presynaptic release is currently unclear, levels of Munc18, Syp, and Syt3, are all increased in +/M2145T mice at P42 and can contribute to the enhanced Pr and altered SV cycling. The unusual PPR in the +/M2145T variant may result, at least in part, from altered Syt3 levels, crucial for calcium-dependent SV replenishment, as its deficiency causes similar phenotypes (128).

Our findings show that Trio GEF1 and GEF2-mediated Rho GTPase signaling pathways play critical and distinct roles in regulating Pr, RRP size and SV recycling.

Conclusions

TRIO is a risk gene for several NDDs with different patterns of variants observed in different disorders. We show here that variants in Trio that lead to impaired Trio levels or GEF function cause both shared and distinct defects in behavior, neuroanatomy, and prominent synaptic dysfunction that may reflect variant-specific NDD clinical phenotypes. Our data also demonstrate, for the first time, the differential impact of distinct Trio lesions on glutamate Pr, RRP size, and SVs replenishment, along with alterations in presynaptic release machinery that contribute to these deficits. We demonstrate that the +/K1431M lesion in TRIO GEF1 leads to presynaptic deficits due to Rac1 dysregulation, which can be rescued by pharmacological normalization. We propose that Trio variants disrupt brain function by impairing Rho GTPase signaling, causing diverse synaptic dysfunction through combined effects on pre-and/or post-synaptic functions. Our findings support growing evidence linking synaptic dysfunction in several NDD models to Rho GTPase signaling dysregulation, identifying it as a commonly affected pathway and a potential therapeutic target.